---
title: "Principal Components Analysis"
subtitle: "High Dimensional Data Analysis"
author: "Anastasios Panagiotelis"
date: "Lecture 3"
output:
  xaringan::moon_reader:
    chakra: libs/remark-latest.min.js
    lib_dir: libs
    css: [default,"mtheme.css","mod.css"]
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    includes:
      after_body: extra.html
    
---

class: inverse, center, middle

# Motivation

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = FALSE, fig.align = "center")
require(magrittr)
require(tidyverse)
require(plotly)
require(widgetframe)
require(animation)
require(scatterplot3d)
require(DT)
require(rgl)
require(knitr)
require(kableExtra)
rm(list=ls())
```

---

# Motivation

- In **marketing** surveys we may ask a large number of questions about customer experience.<!--D-->
--

- In **finance** there may be several ways to assess the credit worthiness of firms.<!--D-->
--

- In **economics** the development of a country or state can be measured in different ways.

---

# A real example

- Consider a dataset with the following variables for the 50 States of the USA<!--D-->
--

  + Income
  + Illiteracy
  + Life Expectancy
  + Murder Rate
  + High School Graduation Rate<!--D-->
--

- You can access this via moodle from the file *StateSE.RData*<!--D-->
--

- Let's do some exploratory data analysis

---

# Life Exp v High School Grad.

```{r levhs}
load('StateSE.RData') #Load Data (check path)
ggplot(StateSE,aes(x=LifeExp,y=HSGrad,text=State))+
  geom_point(size=4)+theme(text=element_text(size=20))->gp#Text made bigger
ggplotly(gp,dynamicTicks = TRUE,tooltip = 'text')%>%
  frameWidget(width="70%",height="40%")
```

---

# Three dimensions

```{r,  testgl, webgl=TRUE, cache=TRUE, echo=FALSE,fig.align='center',warning=FALSE,messages=FALSE,fig.height=8,fig.width=10,eval=FALSE}
library(rgl)
knit_hooks$set(webgl = hook_webgl)
invisible(open3d())
plot3d(x = StateSE$LifeExp, y = StateSE$HSGrad, z = StateSE$Murder, xlab='Life Expectancy', ylab='High School Graduation', zlab='Murder Rate', size=8)
rglwidget()
```

---

# All five variables

- Can not do a scatter plot in five dimensions.<!--D-->
--

- Can we find a single variable to summarise the information in this data set?<!--D-->
--

- Going beyond that can we *rotate* the data to find a good 2-dimensional representation?<!--D-->
--

- Both of these ideas can be achieved using **Principal Components**

---

# Summarising many variables

- Often we aim to  combine many variables into a single index?<!--D-->
--

  + In finance a credit score summarises all the information about the likelihood of bankruptcy for a company.<!--D-->
--

  + In marketing we require a single overall measure of customer experience.<!--D-->
--

  + In economics the Human Development Index is a single measure that takes income, education and health into account.

---

# Weighted linear combination

- A convenient way to combine variables is through a *linear combination* (LC)<!--D-->
--

  + For example, your grade for this unit:
$$ 
w_1\mbox{Assign. Marks}+w_2\mbox{Exam Mark}
$$
  + Here $w_1$ and $w_2$ are called *weights*
  + In this unit, the  weight for the Assignments is *50%* and for the Examination is *50%*<!--D-->
--

- What is a good way to choose weights?

---

# Maximise variance

- The purpose of grading students is to differentiate the best perfoming students from the weakest performing students <!--D-->
--

- The index should have *large variance*.<!--D-->
--

- The LC with the highest variance is the **first Principal Component** of the data.<!--D-->
--

- The first principal component is a new variable that *explains* as much variance as possible in the original variables.

---

# Original Data

```{r original}
kable(StateSE)%>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"))%>%
  scroll_box(height="500px")
```

---

# First PC

```{r, echo=FALSE,message=FALSE}
load('StateSE.RData')
StateSE%>%
  select_if(.,is.numeric)%>%
  scale()%>%
  prcomp()->pca
PCdf<-as.data.frame(pca$x)%>%mutate(.,State=StateSE$State)
StateSEPC<-left_join(StateSE,PCdf)
kable(StateSEPC%>%select(.,State,PC1),format = 'html')%>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"))%>%
  scroll_box(height="500px")
```

---

# First PC on Map

```{r, echo=FALSE, fig.align='center',fig.width=14,fig.height=10,message=FALSE}
state_map<-map_data("state")
ggplot(StateSEPC,aes(map_id=State))+
  coord_map("albers",lat0=30,lat1=40)+
  geom_map(aes(fill=PC1), map=state_map)+
  geom_text(aes(x=state.center$x,y=state.center$y,label=StateAbb),size=9)+
  expand_limits(x=state_map$long, y=state_map$lat)+
  scale_fill_gradient2()+#n(colors=heat.colors(20))+
  theme_dark()+
  labs(x="",y="")
```

---

# Second Principal Component

- Sometimes a single index still oversimplifies the data.<!--D-->
--

- The second principal component is an LC that<!--D-->
--

  1. Is uncorrelated with the first PC.
  2. Has the highest variance out of all LCs that satisfy condition 1.<!--D-->
--

-  Since there is no need for PC2 to *explain* any variance already explained by PC1, PC2 and PC1 are uncorrelated.<!--D-->
--

- We can plot the first two principal components on a scatter plot.

---

# Scatter-plot of PCs

```{r, echo=FALSE, fig.align='center'}
ggplot(StateSEPC,aes(x=PC1,y=PC2, label=StateAbb, text=State))+
  geom_text(size=8)+theme(text= element_text(size=24))->gp
ggplotly(gp,dynamicTicks = TRUE, tooltip = 'text')%>%
  frameWidget(width="70%",height="40%")
```

---

# The weights

```{r weights }
kable(pca$rotation[,1:2],format = 'html')
```
- A high (low) weight indicates a strong positive (negative)
association between a variable and the corresponding PC.

---

# Biplot

- The weight vectors can be plotted on the same scatterplot as
the data.<!--D-->
--

- This is called a biplot.<!--D-->
--

- We can do several useful things with a biplot<!--D-->
--

  + See how the observations relate to one another
  + See how the variables relate to one another
  + See how the observations relate to the variables
  
---

# Types of biplot

- There are multiple ways to draw a biplot.<!--D-->
--

- We will look at two versions<!--D-->
--

  + Distance Biplot
  + Correlation Biplot

---

# Distance Biplot

```{r dbiplot, fig.width=14,fig.height=10}
rownames(pca$x)<-use_series(StateSE,StateAbb)
biplot(pca,cex=2,xlim=c(-0.4,0.3))
```

---

# Distance Biplot

- The distance between observations implies similarity between observations<!--D-->
--

   + Louisiana (LA) and South Carolina (SC) are close therefore are similar.
   + Arkansas (AR) and California (CA) are far apart and therefore different.<!--D-->
--

- If the variables are ignored this is identical to a scatter plot of principal components.

---

# Correlation Biplot

```{r cbiplot, fig.width=14,fig.height=10}
rownames(pca$x)<-use_series(StateSE,StateAbb)
biplot(pca,cex=2,xlim=c(-4.5,2.6),ylim=c(-2,4),scale=0)
```

---

# Correlations

```{r cormat}

StateSE%>%
  select_if(is.numeric)%>%
  cor%>%
  kable(digits = 3)%>%
  kable_styling(font_size=14)

```

---

# Correlation Biplot

- The angles between variables tell us something about correlation (approximately)
  + Income and HSGrad are highly positively correlated.  The angle between them is close to zero.
  + LifeExp and Income are close to uncorrelated.  The angle between them is close 90 degrees.
  + Murder and LifeExp are highly negatively correlated.  The angle between them is close 180 degrees.

---

# More comparison

- The biplot also allows us to compare observations to variables.<!--D-->
--

- Think of the variables as axes.<!--D-->
--

- Draw the shortest line from each point to the axis.<!--D-->
--

- The position along that axis gives an approximation to the actual value of the variable for that observation.

---

# Usual scatter plot

```{r levhs2}
load('StateSE.RData') #Load Data (check path)
ggplot(StateSE,aes(x=LifeExp,y=HSGrad,text=State))+
  geom_point(size=4)+theme(text=element_text(size=20))->gp#Text made bigger
ggplotly(gp,dynamicTicks = TRUE,tooltip = 'text')%>%
  frameWidget(width="70%",height="40%")
```

---

# Biplot

```{r dbiplot2, fig.width=14,fig.height=10}
rownames(pca$x)<-use_series(StateSE,StateAbb)
biplot(pca,cex=2,xlim=c(-0.35,0.25),ylim=c(-0.25,0.25))
```

---

# More PCs

- We can find a third PC, which has the highest variance, while
being uncorrelated with PC1 and PC2.<!--D-->
--

- We cannot visualise this with a biplot, but there are alternatives depending on the structure of the data.<!--D-->
--

- Now a time series example where we consider 3
principal components.

---

# A Time Series Example

- The Stock and Watson dataset contains data on 109
macroeconomic variables in the following categories<!--D-->
--

  + Output
  + Prices
  + Labour
  + Finance<!--D-->
--

- One cannot look at 109 time series plots to visualise general
macroeconomic conditions.<!--D-->
--

- However, one can look at time series plots of the principal
components of these variables.

---

# Plots of PCs

```{r sw}
load('SWMacro.RData')
pcaSW<-prcomp(StockWatsonMacro,scale. = TRUE)
pcaSW%>%
  use_series(x)%>%
  as.data.frame%>%
  rownames_to_column->pcd 
start <- as.Date("1960-01-01")
end <- as.Date("2008-10-01")
sd<-seq.Date(start,end,by='quarters')
mutate(pcd,Time=sd)%>%
  select(.,Time,PC1,PC2,PC3)%>%
  gather(.,key=Comp,value=PC,PC1,PC2,PC3)%>%
  ggplot(.,aes(x=Time,y=PC))+geom_line(size=2)+facet_wrap(~Comp,nrow=3)
```

---

# All PCs

- There are as many principal components as there are variables.
- Together all $p$ principal components explain all of the variation in all $p$ original variables.
$$\sum_{j=1}^p \mbox{Var}(C_j)=\sum_{j=1}^p \mbox{Var}(Y_j)$$
- Where $C_j$ is principal component $j$ and $Y_j$ is variable $j$
---

#So why PCs

- However a small number of principal components can often explain a large proportion of the variance<!--D-->
--

  + In the first example, 2 PCs explain 84% of the total variation of 5 variables.<!--D-->
--

  + In our second example, 3 PCs explain 35% of the total variation of 109 variables.

---
   
# Summary

- Principal components analysis is useful for<!--D-->
--

  + Creating a single index<!--D-->
--

  + Seeing how variables are associated with observations on a single biplot.<!--D-->
--

  + Visualising high-dimensional time series.<!--D-->
--

- How do we do it?

---
class: inverse, middle, center

# Implementation of PCA

---

# Restriction

- Recall that the objective is to find an LC with a large variance.  How could we ‘cheat’ ?<!--D-->
--

  + For a single variable $\mbox{Var}(wY) = w^2 \mbox{Var}(Y)$ <!--D-->
--

  + The variance can be made large by choosing a huge value of $w$.<!--D-->
--

- For this reason the following restriction (normalization) is used
$$w_1^2 + w_2^2 \ldots + w_p^2 = 1.$$

---

# Standardisation

- A similar logic applies to the units that the variables are
measured in.<!--D-->
--

- In the states dataset, income varies from $3000
to $6000, life expectancy varies from 67 years to 73
years.<!--D-->
--

  + Which variable will probably have the larger variance?<!--D-->
--

- Income likely to have a larger variance.

---

# Different units

- If income is measured in \$ ’000s then it will vary from about 3
to 6, If Life Expectancy in measured in days rather than years
it will vary from about 24800 days to 26900 days<!--D-->
--

  + Which variable will have the larger variance now?<!--D-->
--

- The weights can be influenced by the units of measurement.

---

# Effect of standardisation

```{r stdweights}
Std<-pca$rotation[,1]
StateSE%>%
  select_if(is.numeric)%>%
  prcomp%>%
  use_series(rotation)%>%
  magrittr::extract(,1)->Unstd

StateSE%>%
  select_if(.,is.numeric)%>%
  mutate(Income=Income/1000,LifeExp=LifeExp*365.25)%>%
  prcomp%>%
  use_series(rotation)%>%
  magrittr::extract(,1)->DifUnits

kable(cbind(Std,Unstd,DifUnits),digits=4)
```

---

# Standardise or not?

- While the normalisation $w_1^2 + w_2^2+\ldots+ w_p^2 = 1$ is always
implemented in any software that does PCA, the decision to
standardise is up to you.<!--D-->
--

- If the variables are measured in the *same* units then
  + *No* need to standardise.<!--D-->
--

- If the variables are measured in the *different* units then
  + *Standardise* the data.

---
    
# Principal Components in R

- There are several functions for doing Principal Components
Analysis in R. We will use `prcomp`<!--D-->
--

- We can scale in two ways<!--D-->
--

  + Scale the data using the function scale
  + Include the option `scale.=TRUE` when calling the function `prcomp`<!--D-->
--

- Now we will do PCA on the states dataset using R    

---

# Principal Components in R

```{r prcomp, echo=TRUE}
StateSE%>%
  select_if(is.numeric)%>% #Only use numeric variables
  prcomp(scale. = TRUE)->pcaout #Do pca 
summary(pcaout) #summary of information
```  

---

# Principal Components in R

- The output of the `prcomp` function is a prcomp object. <!--D-->
--

- It is a list that contains a lot of information.  Of most interest are<!--D-->
--

  + The principal components which are stored in `x`
  + The weights which are stored in `rotation`


---

# Biplot

- The biplot can be produced by:

```{r bplotnolab,echo=TRUE,eval=FALSE}
biplot(pca)
```

- To have the state abbreviations on the plot they need to be attached to the matrix `pca$x`

```{r biplotlab,echo=TRUE,eval=FALSE}
rownames(pca$x)<-use_series(StateSE,StateAbb)
biplot(pca)
```

- Try it!

---

# Correlation biplot

- By default `biplot` produces the distance biplot.
- To produce the correlation biplot try
```{r, eval=FALSE,echo=TRUE}
biplot(pca,scale = 0)
```

---

# Scree Plot

- Another plot that is easy to create is the Scree plot.<!--D-->
--

- Along the horizontal axis is the Principal Component.<!--D-->
--

- Along the vertical axis is the variance corresponding to each
Principal Component.<!--D-->
--

- The Scree plot indicates how much each PC explains the total
variance of the data.<!--D-->
--

```{r,echo=TRUE,eval=FALSE}
screeplot(pca,type="lines")
```

---

# Scree Plot

```{r,echo=FALSE,eval=TRUE}
screeplot(pca,type="lines")
```

---

# Selecting the number of PCs

- The Scree plot can be used to select the number of Principal
Components.<!--D-->
--

- Look for a part where the plot flattens out also called the
elbow of the Scree Plot.<!--D-->
--

- Another criterion used for standardised data is Kaiser’s Rule.
The rule is to select all PCs with a variance greater than 1.

---

# Number of PCs

- The way PCs are selected depend on the nature of the analysis.<!--D-->
--

- For a visualisation via the biplot, two PCs must be selected.<!--D-->
--

- In this case check the proportion of variance explained by those PCs<!--D-->
--

- The higher this number the more accurate the biplot

---

# Towards Factor Analysis

- For survey data it is often the case that multiple survey questions are measures of the same underlying factor.<!--D-->
--

- For example, at the end of semester you evaluate this unit.  <!--D-->
--

- Typically you will be asked many questions.<!--D-->
--

- This is no different from any other customer satisfaction survey

---

# Underlying factors

- Although you are asked many questions perhaps there are two underlying factors that drive<!--D-->
--

  + The quality of the course materials
  + The quality of the teaching staff<!--D-->
--

- Perhaps the quality of assessment is a third factor.<!--D-->
--

- For survey data, Scree plots and Kaiser's rule can be used to select the number of underlying factors.