---
author: "Department of Econometrics and Business Statistics, Monash University"
date: "Tutorial 7"
output: pdf_document
---

```{r, echo = FALSE}
#rmarkdown::render('MatrixBasics.Rmd',output_file='MatrixBasics.pdf')
#rmarkdown::render('MatrixBasics.Rmd',output_file='MatrixBasicsSols.pdf')
sols<-TRUE
title<-ifelse(sols, 'HDDA Tutorial: MatrixBasics
: Solutions','HDDA Tutorial: MatrixBasics'
)
```

---
title: "`r title`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Vectors

Consider the vectors

$$
\mathbf{a}=\begin{pmatrix}2 \\\ 4 \end{pmatrix} \quad \mathbf{b}=\begin{pmatrix}1 \\\ 0 \end{pmatrix} \quad \mathbf{c}=\begin{pmatrix}-2 \\\ 1 \end{pmatrix} \quad
$$
Work out the following (without using R). All multiplication is matrix multiplication.

1. $\mathbf{a}+\mathbf{b}$
2. $\mathbf{a}'\mathbf{a}$
3. $\mathbf{a}'\mathbf{b}$
4. $\mathbf{a}'\mathbf{c}$
5. $\mathbf{a}\mathbf{b}'$

`r if(sols) '*Solutions provided in section using R*'`

# Matrices

Consider the matrices

Work out the following (without using R). All multiplication is matrix multiplication.

$$
\mathbf{X}=\begin{pmatrix}1& 2 \\\ 1& 4 \\\ 0 & -1 \end{pmatrix} \quad \mathbf{Y}=\begin{pmatrix}2 & -1 \\\ 3 & 0 \\\ 3 & -1 \end{pmatrix}
$$

1. $\mathbf{X}+\mathbf{Y}$
2. $\mathbf{X}\mathbf{Y}$
3. $\mathbf{X}'\mathbf{Y}$

`r if(sols) '*Solutions provided in section using R*'`

# Vectors and Matrices in R

Repeat all above questions using R.  Useful functions are `c` for setting a vector, `matrix` for setting a matrix and `t` for the transpose.  Also note that `*` does NOT do matrix multiplication.  Instead use `%*%`.

```{r, eval=sols,echo=sols,message=FALSE}
#Vectors

a<-c(2,4)
b<-c(1,0)
c<-c(-2,1)

a+b
t(a)%*%a
t(a)%*%b
t(a)%*%c
a%*%t(b)

#Matrices

#Make sure you set byrow=T
X<-matrix(c(1,2,
            1,4,
            0,-1),3,2,byrow = T)
X

Y<-matrix(c(2,-1,
            3,0,
            3,-1),3,2,byrow = T)
Y

X+Y
# X%*%Y This is non-conformable
t(X)%*%Y
```

# Data matrix

Consider the data matrix $\mathbf{Y}$

$$
\mathbf{Y} = \begin{pmatrix}y_{11} & y_{12} & \ldots & y_{1p} \\\ y_{21} & y_{22} & \ldots & y_{2p} \\\ \vdots & \vdots &\ldots &\vdots \\\ y_{n1} & y_{n2} & \ldots & y_{np} \end{pmatrix}
$$

where $y_{ij}$ is the value of variable $j$ for observation $i$ 

1. How many rows are there in $\mathbf{Y}$?

`r if(sols) '*There are n rows, i.e. each observation is a row*'`

2. How many columns are there in $\mathbf{Y}$?

`r if(sols) '*There are p columns, i.e. each variable is a column*'`

3. What are the dimensions of $\mathbf{Y}$?

`r if(sols) '*The matrix is an $n\\times p$ matrix*'`

4. Find an expression for the first row and first column of $\mathbf{S}=\frac{1}{n-1}\mathbf{Y}'\mathbf{Y}$

`r if(sols) '*The element on the first row and first column is found by multiplying $\\begin{pmatrix}y_{11} & y_{21} & \\ldots & y_{n1}\\end{pmatrix}$ by its transpose. This is the same as $y^2_{11}+y^2_{21}+\\ldots+y^2_{n1}$.  Putting everything together it is $\\frac{1}{n-1}\\sum\\limits_{i=1}^{n}y^2_{i1}$.  If the data have mean zero, then this is the sample variance of the first variable.*'`
