---
author: "Department of Econometrics and Business Statistics, Monash University"
date: "Tutorial 2"
output: pdf_document
---

```{r, echo = FALSE}
#rmarkdown::render('IntroR.Rmd',output_file='IntroR.pdf')
#rmarkdown::render('IntroR.Rmd',output_file='IntroRSols.pdf')
sols<-TRUE
title<-ifelse(sols, 'HDDA Tutorial: Getting Started with R
: Solutions','HDDA Tutorial: Getting Started with R'
)
```

---
title: "`r title`"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The aim of this week's tutorial is to do more preliminary data analysis using R.  You will need to use the datasets *Beer.rds* and *comScore.rds* from which can be downloaded from Moodle.

## Beer Data

1. Without using the `qplot` function, produce a histogram of the cost per 12 fl. oz. variable.

```{r, eval=sols,echo=sols,message=FALSE}
#First load required packages
library(dplyr)
library(ggplot2)
#Then load in data
Beer<-readRDS('Beer.rds')
#Histogram of cost
ggplot(Beer,aes(x=cost))+geom_histogram()
# In the histogram notice the presence of the outlier.
```

2. Without using the `qplot` function, produce boxplots of alcohol content.  On the same plot there should be a separate boxplot for light beers and a separate boxplot for nonlight beers.

```{r, eval=sols,echo=sols,message=FALSE}
#Boxplots
ggplot(Beer,aes(x=light,y=alcohol))+geom_boxplot()
# Unsuprisingly the light beers have less alcohol content than the 
#non-light beers.
```

3. Produce a frequency table of beer rating

```{r, eval=sols,echo=sols,message=FALSE}
#Boxplots
table(Beer$rating)
# There are roughly equal numbers of fair good and very good beers.
```

4. Produce a cross tab of beer rating against light/nonlight

```{r, eval=sols,echo=sols,message=FALSE}
#Boxplots
table(Beer$rating,Beer$light)
# Notice that zero light beers are rated very good.  This suggests 
#a relationship between light/non-light and the rating of beer.
```

## comScore Data

The company comScore records the online behaviour of subscribers.  Each observation of the dataset that you have been provided with is a unique visit to the wesite apple.com.  Four variables are recorded: **Buy** indicates whether a purchase was made, **Sales** indicates the value of any purchase, **Duration** indicates how much time was spent on apple.com, while **PageViews** indicates how many pages were clicked on under the apple.com domain name in a single visit.  An interesting marketing question is whether browsing behaviour (duration and page views) are associated with purchase behaviour (buy and sales).

1. Without using the `qplot` function, produce histograms of
  - Sales
  - Duration
  - Page Views

```{r, eval=sols,echo=sols,message=FALSE}
#Then load in data
comScore<-readRDS('comScore.rds')
#Histogram of sales
ggplot(comScore,aes(x=Sales))+geom_histogram()
#Histogram of duration
ggplot(comScore,aes(x=Duration))+geom_histogram()
#Histogram of page views
ggplot(comScore,aes(x=PageViews))+geom_histogram()

#The resulting histograms are not particularly informative since 
#low values are so prevalent and since the data is very highly 
#skewed.
```  
  
2. Produce summary statistics for the comScore data

```{r, eval=sols,echo=sols,message=FALSE}
summary(comScore)
#Notice that most visits do not result in a purchase.  For this 
#reason means are higher than medians. Also the median value of 
#Sales is 0.  Finally, notice that the maximum values of Sales, 
#Duration and Page Views are all very high.
```

3. Without using the `qplot` function, produce a scatter plot of duration against page views

```{r, eval=sols,echo=sols,message=FALSE}
ggplot(comScore,aes(x=Duration,y=PageViews))+geom_point()
# There are two obvious outliers.  The outlier in page views 
#implies that 1071 clicks were recorded in 2 minutes.  This may be 
#due to an error in how page views are recorded or a bot.  The 
#outlier in duration implies that 64 pages were clicked on in over 
#23 hours.  Here it is possible that the user simply left the 
#computer on.
```

## More Advanced

1. Create a new data frame that removes the two outliers using the `filter` function in `dplyr`.

```{r, eval=sols,echo=sols,message=FALSE}
# Only select Page Views and Duration
# that are less than maximum values
comScore_no<-filter(comScore,
                    (PageViews<1071)&
                      (Duration<1428))

#Note you must use the and operator. As an additional exercise can 
#you think of another way to complete this task?
```

2. Do the scatterplot again with the outliers removed.

```{r, eval=sols,echo=sols,message=FALSE}
# Same as before, simply change the data frame
ggplot(comScore_no,aes(x=Duration,y=PageViews))+geom_point()

#There is a positive relationship between page views and clicks
#however the points almost appear to come from two different 
#distributions, one where the correlation is stronger than the  
#other
```

3. Do the scatterplot where the points have a different colour if the observation corresponds to a buy and a different colour if it corresponds to no buy.

```{r, eval=sols,echo=sols,message=FALSE}
#Add the color aesthetic
ggplot(comScore_no,aes(x=Duration,y=PageViews,col=Buy))+
  geom_point()

#Here the group with low correlation between page views and #duration are seen to mostly be non-purchasers.  The group where
#the correlation is higher has a higher incidence of buy.  This
#may suggest that websites that are sticky, i.e. involve more #interaction result in a higher purchase probability.  #Alternatively, the process of buying a product on a website may 
#involve multiple page views.

```