---
title: "Dimension Reduction"
subtitle: "High Dimensional Data Analysis"
author: "Anastasios Panagiotelis"
date: "Lecture 10"
output:
  xaringan::moon_reader:
    chakra: libs/remark-latest.min.js
    lib_dir: libs
    css: [default,"mtheme.css","mod.css"]
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false

---

class: inverse, center, middle

# A proof

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo=FALSE , fig.align = "center")
require(magrittr)
require(tidyverse)
require(plotly)
require(widgetframe)
require(animation)
require(scatterplot3d)
require(DT)
require(rgl)
require(knitr)
require(kableExtra)
require(animation)
```

---

# How to do PCA

- Over the next few slides, we will derive how to get the first principal component.
--

- This eventually leads us to the *eigenvalue decomposition*.
--

- The eigenvalue decomposition and its more general form known as the *singular value decomposition* are crucial to all dimension reduction techniques.
--

- This will be challenging.

---

# First PC

- Recall that the principal component is 
  - The linear combination of the variables
  - With maximum variance
  - Subject to the squared weights summing to 1.

---

# Linear Combination

- Let $c_i=w_1y_1+\ldots+w_py_p$ 
--

- In matrix/vector form $c_i=\mathbf{w}'\mathbf{y}_i$ 
--

- What are the dimensions of $c_i$, $\mathbf{w}$ and $\mathbf{y}_i$?
--

- Both $\mathbf{w}$ and $\mathbf{y}_i$ are $p\times 1$ vectors, while $c_i$ is a scalar.

---

# Variance

- The variance of the linear combination is given by

$$\mbox{Var}(c)=\frac{\sum\limits_{i=1}^n c^2_i}{n-1} =\frac{\sum\limits_{i=1}^n(\mathbf{w}'\mathbf{y}_i)^2}{n-1}$$
- Assumed that all $y$'s have a mean of zero.  
- Everything still works without this assumption but is messier.

---

# A trick

- Note that $w_1y_1+\ldots+w_py_p$ has been written as  $\mathbf{w}'\mathbf{y}_i$, but...
--

- ... it can also be written as $\mathbf{y}'_i\mathbf{w}$.
--

- This in turn implies that $c^2_i=\mathbf{w}'\mathbf{y}_i\mathbf{y}_i'\mathbf{w}$.  Substituting into the variance formula gives.
--

$$\mbox{Var}(c)=\frac{\sum\limits_{i=1}^n c^2_i}{n-1} = \frac{\sum\limits_{i=1}^n\mathbf{w}'\mathbf{y}_i\mathbf{y}_i'\mathbf{w}}{n-1}$$

---

# Linearity

Linearity implies that anything without an $i$ subscript can be taken outside the summation sign.

$$\mbox{Var}(c)=\frac{\sum\limits_{i=1}^n c^2_i}{n-1} = \frac{\mathbf{w}'\left(\sum\limits_{i=1}^n\mathbf{y}_i\mathbf{y}_i'\right)\mathbf{w}}{n-1}$$

---

#Scalar multiplication

The order of scalar multiplication does not matter allowing the following

$$\begin{align}\mbox{Var}(c)&= \mathbf{w}'\left(\frac{\sum\limits_{i=1}^n\mathbf{y}_i\mathbf{y}_i'}{n-1}\right)\mathbf{w}\\ &= \mathbf{w}'\mathbf{S}\mathbf{w} \end{align}$$

Recall $\mathbf{S}$ is the variance covariance matrix

---

# Objective

We want to choose $\mathbf{w}$ to maximise the variance while ensuring that $w_1^2+\ldots+w_p^2=\mathbf{w}'\mathbf{w}=1$.  We write this as

$$
\begin{align}\underset{\mathbf{w}}{\max}\quad&\mathbf{w}'\mathbf{S}\mathbf{w}\\\ \mbox{s.t.}\quad & \mathbf{w}'\mathbf{w}=1\end{align}
$$

---
class: inverse, center, middle

# Optimisation

---
# Constrained Optimisation

Solving the constrained optimisation above is equivalent to solving the following unconstrained problem

$$\underset{\mathbf{w},\lambda}{\max}\quad\mathbf{w}'\mathbf{S}\mathbf{w}-\lambda(\mathbf{w}'\mathbf{w}-1)$$

---

# Gradient 

- Many optimisation problems involve using the *gradient* or slope of an objective function.
--

- Think of an analogy of climbing a hill.
--

- If the gradient is positive then you can go higher by walking forwards.
--

- If the gradient is negative then you can go higher by walking backwards.
--

- The top of the hill is where the gradient is zero.

---

# Gradient

- To compute the gradient we need to use matrix calculus.
--

- What does it mean to differentiate with respect to $\mathbf{w}$?
--

- It means we differentiate with respect to $w_1$, $w_2$, etc
--

- All up $p$ first derivatives are found.  These can be stored in a vector.

---

# First Order Conditions

Differentiating w.r.t. $\mathbf{w}$ gives

$$\frac{\partial\left(\mathbf{w}'\mathbf{S}\mathbf{w}-\lambda(\mathbf{w}'\mathbf{w}-1)\right)}{\partial \mathbf{w}}=2\mathbf{S}\mathbf{w}-2\lambda{\mathbf{w}}$$

Differentiating w.r.t. $\lambda$ gives

$$\frac{\partial\left(\mathbf{w}'\mathbf{S}\mathbf{w}-\lambda(\mathbf{w}'\mathbf{w}-1)\right)}{\partial \lambda}=\mathbf{w}'\mathbf{w}-1$$
---

# How did we do that?

The key result is that for any square matrix $\mathbf{A}$ it holds that

$$\frac{\partial \mathbf{w}'\mathbf{A}\mathbf{w}}{\partial\mathbf{w}}=2\mathbf{A}\mathbf{w}$$
This is the matrix version of the rule that the derivative of $\partial{aw^2}/\partial w = 2aw$.  From this result, the matrix result can be derived (but this is tedious).

---

# Eigenvalue Problem

The gradient will be zero when $2\mathbf{S}\mathbf{w}-2\lambda{\mathbf{w}}=\mathbf{0}$ or simplifying when

$$ \mathbf{S}\mathbf{w}=\lambda{\mathbf{w}} $$

This is a very famous problem known as the **eigenvalue** problem. Suppose $\tilde{\lambda}$ and $\tilde{\mathbf{w}}$ provide a solutions then
--

- The value of  $\tilde{\lambda}$ is called an *eigenvalue*
- The vector $\tilde{\mathbf{w}}$ is called an *eigenvector*

---

# Eigenvalue Problem

- For $2\times 2$, $3\times 3$ and $4\times 4$ matrices there are formulas for $\tilde{\lambda}$.
--

- These are hideous
--

- For $5\times 5$ and beyond there is no formula
--

- A solution is found using numerical methods (i.e. a computer algorithm).

---

# Geometric View

- Recall that multiplying by a matrix moves vectors around, changing their length and direction.
--

- However for any matrix there will be some vector whose direction does not change, but only the length.
--

- This vector is an eigenvector.
--

- The extent to which the length is changed is the eigenvalue.

---

# Multiple solutions

- In general there are multiple solutions to the eigenvalue problem.
--

- Which one maximises the variance?
--

- Let $\tilde{\mathbf{w}}$ be an eigenvector and its associated eigenvalue be $\tilde{\lambda}$.
--

- What is the variance of the linear combination $\tilde{\mathbf{w}}'\mathbf{y}$?

---

# Answer

We have already shown that the variance will be $\tilde{\mathbf{w}}'\mathbf{S}\tilde{\mathbf{w}}$.  Since $\tilde{\mathbf{w}}$ is an eigenvector it must hold that

$$
\mathbf{S}\tilde{\mathbf{w}}=\tilde{\lambda}\tilde{\mathbf{w}}
$$
--

which implies

$$
\begin{align}\tilde{\mathbf{w}}'\mathbf{S}\tilde{\mathbf{w}}&=\tilde{\mathbf{w}}'\tilde{\lambda}\tilde{\mathbf{w}}\\\ &= \tilde{\lambda}\tilde{\mathbf{w}}'\tilde{\mathbf{w}}&\end{align}
$$

---

# Variance

- Since $\tilde{\mathbf{w}}'\tilde{\mathbf{w}}=1$ this implies that the variance of the linear combination is $\tilde{\lambda}$.
--

- The weights for the first principal component is given by the eigenvector that corresponds to the **largest eigenvalue**. 
--

- The weights of the remaining principal components are given by the other eigenvectors.


---
class: inverse, middle, center

# Matrix Decompositions

---

# Spectral Theorem

Since $\mathbf{S}$ is a symmetric matrix it can decomposed as

$$\underset{(p\times p)}{\mathbf{S}}=\underset{(p\times p)}{\mathbf{W}}\underset{(p\times p)}{\boldsymbol{\Lambda}}\underset{(p\times p)}{\mathbf{W}}'$$
--

- The columns of $\mathbf{W}$ are eigenvectors of $\mathbf{S}$
--

- $\boldsymbol{\Lambda}$ is a matrix with the eigenvalues along the main diagonal and zeros on the off diagonal.
--

- The eigenvalues and eigenvectors can be rearranged so by convention eigenvalues in $\boldsymbol{\Lambda}$ are sorted from largest to smallest.

---

# Rotation

- The full vector of principal components for observation $i$ is given by $\mathbf{c}_i=\mathbf{W}'\mathbf{y}_i$
--

- The eigenvectors are also orthogonal (A proof of why this is true can be provided for anyone who is curious).
--

- Orthogonality implies that the matrix of eigenvectors $\mathbf{W}$ is a *rotation* matrix.
--

- For this reason we consider PCA to be a rotation of the data.

---

# PCA as an approximation

It can be shown that an equivalent way of writing the eigenvalue decomposition is

$$\begin{align}\underset{(p\times p)}{\mathbf{S}}&=\underset{(p\times p)}{\mathbf{W}}\underset{(p\times p)}{\boldsymbol{\Lambda}}\underset{(p\times p)}{\mathbf{W}}'\\\ &=\sum\limits_{j=1}^p\underset{(1\times 1)}{\lambda_j}\underset{(p\times 1)}{\mathbf{w}_j}\underset{(1\times p)}{\mathbf{w}_j'}\end{align}$$

---
# PCA as an approximation

If some eigenvalues are small they can be ignored.

$$\begin{align}\mathbf{S}& =\sum\limits_{j=1}^p{\lambda_j}{\mathbf{w}_j}{\mathbf{w}_j'}\\\ &\approx  \sum\limits_{j=1}^r{\lambda_j}{\mathbf{w}_j}{\mathbf{w}_j'}\end{align}$$

Only $r<<p$ eigenvalues are used.

---

# Decomposition

- Consider a $50\times 50$ covariance matrix.
--

- There are 1275 variances and covariances to estimate
--

- Suppose the data can be summarised by just 5 factors/principal components.
--

- Then the matrix can be approximated with just 5 eigenvalues and eigenvectors (255 numbers).

---

# In General

- For matrix $\mathbf{X}$ that is possibly non-symmetric and possibly non-square a similar decomposition known as the singular value decomposition can be used.

$$
\underset{(n\times p)}{\mathbf{Y}}=\underset{(n\times n)}{\mathbf{U}}\underset{(n\times p)}{\mathbf{D}}\underset{(p\times p)}{\mathbf{V}'}
$$

The matrices $\mathbf{U}$ and $\mathbf{V}$ are rotations

---

# Structure of D

- If $n>p$
$$\begin{bmatrix}d_1 &\cdots &0\\\vdots &\ddots &\vdots\\0&\cdots&d_p\\0 &\cdots &0\\\vdots &\vdots &\vdots\\0&\cdots&0\end{bmatrix}$$
---

# Structure of D

- If $n<p$
$$\begin{bmatrix}d_1 &\cdots &0 &0 &\cdots &0\\\vdots &\ddots &\vdots&\vdots &\vdots &\vdots\\0&\cdots&d_n&0 &\cdots &0\end{bmatrix}$$
- In both cases all $d_i>0$
--

- These are called *singular values*.

---

# Singular Values

- The singular values are ordered from largest to smallest allowing for an approximation

$$\begin{align}\mathbf{Y}& =\sum\limits_{j=1}^p{d_j}{\mathbf{u}_j}{\mathbf{v}_j'}\\\ &\approx  \sum\limits_{j=1}^r{d_j}{\mathbf{u}_j}{\mathbf{v}_j'}\end{align}$$

for $r<<min(n,p)$

---

# Biplots and the SVD

- When $\mathbf{Y}$ is the data matrix there is a connection between the biplot and the SVD.
--

- For the distance biplot, the first two columns of ${\mathbf U}{\mathbf D}$ are plotted as points and the first two columns of ${\mathbf V}$ as arrows
--

- For the correlation biplot plot the first two columns of ${\mathbf U}$ are plotted as points the first two columns of ${\mathbf V}{\mathbf D}$ as arrows
--

- In general plot the first two columns of ${\mathbf U}{\mathbf D}^\kappa$ and the first two columns of ${\mathbf V}{\mathbf D}^{(1-\kappa)}$<!--D-->
--

- In R, $\kappa$ is set by the scale option of `biplot`

---
class: inverse, center, middle

# A final example

---

# A picture

```{r,echo=FALSE,warning=FALSE}
library(pixmap)
a<-read.pnm('./cat.pgm')
plot(a)

```

---

# Pixels

- For the computer this picture is a matrix<!--D-->
--

- Each pixel on the screen has a number between 0 and 1.<!--D-->
--

  + Numbers closer to 0 display as lighter shades of grey<!--D-->
--

  + Numbers closer to 1 display as darker shades of grey<!--D-->
--

- What if we do the SVD on this matrix?

---

# SVD

- All up there are $232\times 218=50576$ pixels.<!--D-->
--

- Suppose we approximate this matrix with 20 singular values<!--D-->
--

- Then ${\mathbf U_{(r)}}$ is $232\times 20=4640$<!--D-->
--

- Then ${\mathbf V_{(r)}}$ is $218\times 20=4360$<!--D-->
--

- Including the 20 singular values themselves, we summarise 50576 numbers using only $4640+4360+20=9020$ numbers.

---

# Approximation

```{r,echo=FALSE}

mat<-a@grey
mat<-(mat*0.999)+0.0005
matl<-log(mat/(1-mat))

svdmat<-svd(matl)
r<-20
matapprox<-svdmat$u[,1:r]%*%diag(svdmat$d[1:r])%*%t(svdmat$v[,1:r])
#matld<-svdmat$u[,(r+1):218]%*%diag(svdmat$d[(r+1):218])%*%t(svdmat$v[,(r+1):218])
#matapprox[matapprox<0]<-0
#matapprox[matapprox>1]<-1

matapproxl<-1/(1+exp(-matapprox))
#matapproxl<-(matapproxl-0.0005)/0.999
a@grey<-matapproxl
plot(a)

```

---

# Discussion

- Using only 20 singular values we do not lose much information.<!--D-->
--

- What if we reconstruct the picture using singular value 21 to singular value 218?<!--D-->
--

- This uses a lot more information.  Does it give a clearer approximation?

---

# Using remaining singular values

```{r, echo=FALSE}

#matapprox<-svdmat$u[,1:r]%*%diag(svdmat$d[1:r])%*%t(svdmat$v[,1:r])
matapprox<-svdmat$u[,(r+1):218]%*%diag(svdmat$d[(r+1):218])%*%t(svdmat$v[,(r+1):218])
#matapprox[matapprox<0]<-0
#matapprox[matapprox>1]<-1

matapproxl<-1/(1+exp(-matapprox))
a@grey<-matapproxl
plot(a)

```

---

# Singular values

```{r, echo=FALSE}
plot(1:218,svdmat$d,"l",xlab='Singular Value',ylab='',lwd=4)
```

---

# Conclusion

- The main idea is that the SVD summarises the *important* information in the matrix into a small number of singular values.<!--D-->
--

- Rotating so that we can isolate the dimensions associated with those singular values is the geometry behind dimension reduction.<!--D-->
--

- This applies to PCA, factor analysis and MDS as well as to compressing images.