---
title: "Geometry of Matrices"
subtitle: "High Dimensional Data Analysis"
author: "Anastasios Panagiotelis"
date: "Lecture 9"
output:
  xaringan::moon_reader:
    chakra: libs/remark-latest.min.js
    lib_dir: libs
    css: [default,"mtheme.css","mod.css"]
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false

---

class: inverse, center, middle

# Why?

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo=FALSE , fig.align = "center")
require(magrittr)
require(tidyverse)
require(plotly)
require(widgetframe)
require(animation)
require(scatterplot3d)
require(DT)
require(rgl)
require(knitr)
require(kableExtra)
require(animation)
```

---
# Why vectors and matrices?

- Nearly all the methods we cover involve *vectors* and *matrices*.<!--D-->
--

- The earlier lecture on matrices may have felt like it involved remembering many seemingly arbitrary rules.<!--D-->
--

- The aim of this lecture is to understand the *geometric* interpretation of vectors which is crucial in making the connection between data, data analysis and visualisation.<!--D-->
--

- Matrices are important to gain a deep understanding about dimension reduction.

---

# Vectors

- We will think of a 2-dimensional vector in one of three ways:<!--D-->
--

  + A point in 2-dimensional space<!--D-->
--

  + An arrow pointing to that point in 2-dimensional space<!--D-->
--

  + An array of 2 numbers<!--D-->
--

- All of these ideas work for n-dimensional vectors where n can be 3,4 or any number
--

- By convention all vectors in this lecture are column vectors.

---

# Vectors as points

```{r points,  fig.align='center',echo=FALSE}
v1<-c(1,2)
v2<-c(3,1)
plot(0,0,'n',xlim=c(-5,5),ylim=c(-5,5),xlab='',ylab='')
lines(c(-5,5),c(0,0),lty=2)
lines(c(0,0),c(-5,5),lty=2)
points(v1[1],v1[2],pch=19,cex=2)
text(v1[1],v1[2],expression(v[1]),pos=4,cex=2)
```

---

#Vectors as points

```{r pointsb,  fig.align='center',echo=FALSE}
v1<-c(1,2)
v2<-c(3,1)
plot(0,0,'n',xlim=c(-5,5),ylim=c(-5,5),xlab='',ylab='')
lines(c(-5,5),c(0,0),lty=2)
lines(c(0,0),c(-5,5),lty=2)
points(v1[1],v1[2],pch=19,cex=2)
text(v1[1],v1[2],expression(v[1]),pos=4,cex=2)
points(v2[1],v2[2],pch=19,cex=2)
text(v2[1],v2[2],expression(v[2]),pos=4,cex=2)
```

---

# Vectors as arrows

```{r arrows,  fig.align='center',echo=FALSE}
v1<-c(1,2)
v2<-c(3,1)
plot(0,0,'n',xlim=c(-5,5),ylim=c(-5,5),xlab='',ylab='')
lines(c(-5,5),c(0,0),lty=2)
lines(c(0,0),c(-5,5),lty=2)
arrows(0,0, v1[1],v1[2],pch=19,lwd=2)
text(v1[1],v1[2],expression(v[1]),pos=4,cex=2)
```
---

# Vectors as arrows

```{r arrowsb,  fig.align='center',echo=FALSE}
v1<-c(1,2)
v2<-c(3,1)
plot(0,0,'n',xlim=c(-5,5),ylim=c(-5,5),xlab='',ylab='')
lines(c(-5,5),c(0,0),lty=2)
lines(c(0,0),c(-5,5),lty=2)
arrows(0,0, v1[1],v1[2],pch=19,lwd=2)
text(v1[1],v1[2],expression(v[1]),pos=4,cex=2)
arrows(0,0, v2[1],v2[2],pch=19,lwd=2)
text(v2[1],v2[2],expression(v[2]),pos=4,cex=2)
```

---

# Vectors as numbers

In general a vector is:

$${\mathbf{v}}=\begin{bmatrix}x\\y\end{bmatrix}$$

For our two examples.

$${\mathbf{v}}_1=\begin{bmatrix}1\\2\end{bmatrix}\quad\mbox{and}\quad{\mathbf{v}}_2=\begin{bmatrix}3\\1\end{bmatrix}$$

---

# Data as a vector

- Consider two examples:<!--D-->
--

  + The vector $\bf{v}$ corresponds to a cage where $x$ measures the number of chickens and $y$ measures the number of rabbits.<!--D-->
--

  + The vector $\bf{v}$ corresponds to a country where $x$ measures the trade deficit in 2016 and $y$ measures trade deficit in 2017.<!--D-->
--

- Notice that if we have many cages or countries and plot each of these vectors as points we get a scatterplot

---

# Vectors as points

```{r,  fig.align='center',echo=FALSE}
v1<-c(1,2)
v2<-c(3,1)
plot(0,0,'n',xlim=c(-5,5),ylim=c(-5,5),xlab='',ylab='')
lines(c(-5,5),c(0,0),lty=2)
lines(c(0,0),c(-5,5),lty=2)
points(v1[1],v1[2],pch=19,cex=2)
text(v1[1],v1[2],expression(v[1]),pos=4,cex=2)
points(v2[1],v2[2],pch=19,cex=2)
text(v2[1],v2[2],expression(v[2]),pos=4,cex=2)
```

---

# Vector Operations

- There are two fundamental *operations* that we can do with vectors<!--D-->
--

  + Vector *addition*<!--D-->
--

  + Scalar *multiplication*<!--D-->
--

- We will look at each of these: <!--D-->
--

  + Geometrically (with pictures) and <!--D-->
--

  + Algebraically (with numbers)

---

#Vector Addition

```{r addition,fig.show='animate',aniopts='controls',interval=0.5,fig.cap='Click to see',ffmpeg.format='ogg'}
v1<-c(1,2)
v2<-c(3,1)
s<-seq(0,1,by=0.2)
for (j in 1:length(s)){
  plot(0,0,'n',xlim=c(-5,6),ylim=c(-5,5),xlab='',ylab='')
  lines(c(-5,5),c(0,0),lty=2)
  lines(c(0,0),c(-5,5),lty=2)
  sv<-s[j]*v2
  vsv<-v1+sv
  arrows(sv[1],sv[2], vsv[1],vsv[2],pch=19,lwd=2,col='blue')
  arrows(0,0, v1[1],v1[2],pch=19,lwd=2)
  text(v1[1],v1[2],expression(v[1]),pos=4,cex=2)
  arrows(0,0, v2[1],v2[2],pch=19,lwd=2)
  text(v2[1],v2[2],expression(v[2]),pos=4,cex=2)
  
  if(j==length(s)){
    arrows(0,0, vsv[1],vsv[2],pch=19,lwd=2,col='red')
    text(vsv[1],vsv[2],expression(v[1]+v[2]),pos=4,cex=2)
  }
}
```
---
  
# The other way 
  
```{r addition2,  fig.align='center',echo=FALSE,fig.show='animate',interval=0.5, label=FALSE,aniopts='controls',fig.cap='Click to see',ffmpeg.format='ogg'}
v1<-c(1,2)
v2<-c(3,1)
s<-seq(0,1,by=0.2)
for (j in 1:length(s)){
  plot(0,0,'n',xlim=c(-5,6),ylim=c(-5,5),xlab='',ylab='')
  lines(c(-5,5),c(0,0),lty=2)
  lines(c(0,0),c(-5,5),lty=2)
  
  sv<-s[j]*v1
  vsv<-v2+sv
  arrows(sv[1],sv[2], vsv[1],vsv[2],pch=19,lwd=2,col='blue')
  
  arrows(0,0, v1[1],v1[2],pch=19,lwd=2)
  text(v1[1],v1[2],expression(v[1]),pos=4,cex=2)
  arrows(0,0, v2[1],v2[2],pch=19,lwd=2)
  text(v2[1],v2[2],expression(v[2]),pos=4,cex=2)
  
  arrows(0,0, 4,3,pch=19,lwd=2,col='red')
  text(4,3,expression(v[1]+v[2]),pos=4,cex=2)
  
}

```

---

  
# Vector Addition: Geometry
  
- The operation of adding two vectors is simply to slide one along the other.<!--D-->
--

- This is called *translation*<!--D-->
--

- The order of adding vectors does not matter $\mathbf{v}_1+\mathbf{v}_2=\mathbf{v}_2+\mathbf{v}_1$<!--D-->
--

- This property is called *commutativity*


---
  
# With lots of points
  
```{r addpts, fig.align='center', echo=FALSE, fig.show='animate', interval=0.5, aniopts='controls', fig.cap='Click to see', ffmpeg.format='ogg'}
grid2<-rbind(rnorm(50),rnorm(50))
v1<-c(1,2)
s<-seq(0,1,by=0.2)
for (j in 1:length(s)){
  plot(0,0,'n',xlim=c(-5.1,5.1),ylim=c(-5.1,5.1),xlab='',ylab='')
  lines(c(-5.1,5.1),c(0,0),lty=2)
  lines(c(0,0),c(-5.1,5.1),lty=2)
  arrows(0,0, v1[1],v1[2],pch=19,lwd=2,col='red')
  text(v1[1],v1[2],expression(v[1]),pos=4,cex=2,col='red')
  sv<-s[j]*v1%*%matrix(rep(1,dim(grid2)[2]),1,dim(grid2)[2])
  vsv<-grid2+sv
  points(vsv[1,],vsv[2,],pch=19,lwd=2,col='blue')
}
```  


---
  
# Vector Additions: Numbers
  
Vector addition is as simple as adding up the corresponding numbers

$$\begin{bmatrix}\color{red}{1}\\\color{black}{2}\end{bmatrix}+
  \begin{bmatrix}\color{red}{3}\\\color{black}{1}\end{bmatrix}=
  \begin{bmatrix}\color{red}{1+3}\\\color{white}{2+1}\end{bmatrix}=
  \begin{bmatrix}\color{white}{4}\\\color{white}{3}\end{bmatrix}$$
  
---
  
# Vector Additions: Numbers
  
Vector addition is as simple as adding up the corresponding numbers

$$\begin{bmatrix}\color{red}{1}\\\color{black}{2}\end{bmatrix}+
  \begin{bmatrix}\color{red}{3}\\\color{black}{1}\end{bmatrix}=
  \begin{bmatrix}\color{red}{1+3}\\\color{white}{2+1}\end{bmatrix}=
  \begin{bmatrix}\color{red}{4}\\\color{white}{3}\end{bmatrix}$$
  
---
  
# Vector Additions: Numbers
  
Vector addition is as simple as adding up the corresponding numbers

$$\begin{bmatrix}\color{black}{1}\\\color{blue}{2}\end{bmatrix}+
  \begin{bmatrix}\color{black}{3}\\\color{blue}{1}\end{bmatrix}=
  \begin{bmatrix}\color{black}{1+3}\\\color{blue}{2+1}\end{bmatrix}=
  \begin{bmatrix}\color{black}{4}\\\color{white}{3}\end{bmatrix}$$
  
---
  
# Vector Additions: Numbers
  
Vector addition is as simple as adding up the corresponding numbers

$$\begin{bmatrix}\color{black}{1}\\\color{blue}{2}\end{bmatrix}+
  \begin{bmatrix}\color{black}{3}\\\color{blue}{1}\end{bmatrix}=
  \begin{bmatrix}\color{black}{1+3}\\\color{blue}{2+1}\end{bmatrix}=
  \begin{bmatrix}\color{black}{4}\\\color{blue}{3}\end{bmatrix}$$
  
---
  
# Vector Addition and Data
  
- Vector addition can be meaningful for data.<!--D-->
--

- If I *add* a cage with 1 chicken and 3 rabbits to another cage with 2 chickens and 1 rabbits what do I get?<!--D-->
--

  + A cage with 3 chickens and 4 rabbits.
---
  
# Vector Addition and Data
  
- Suppose Australia had a trade deficit of $US1bn in 2016 and $US3bn in 2017 and New Zealand had a trade deficit of $US2bn in 2016 and $US1bn in 2017.<!--D-->
--

- Australia and New Zealand merge to form a currency union and the trade deficit statistics need to be updated.<!--D-->
--

  + The new currency union had a trade deficit of $US3bn in 2016 and $US4bn in 2017.


---
  
# Scalar multiplication
  
- If we add a number $c$ times we call that *multiplication*<!--D-->
--

- We do something similar with vectors.<!--D-->
--

- If we add a vector to itself $c$ times this is called *scalar multiplication*.<!--D-->
--

- We are multiplying a vector by a single number and NOT by another vector.<!--D-->
--

- This is very different from *matrix multiplication*.

---
  
# Scalar multiply by 2
  
```{r smultgeom, fig.align='center', echo=FALSE, fig.show='animate', interval=0.5,aniopts='controls', fig.cap='Click to see', ffmpeg.format='ogg'}
v1<-c(1,2)
s<-seq(1,2,0.2)
for (i in 1:length(s)){
  plot(0,0,'n',xlim=c(-5,5),ylim=c(-5,5),xlab='',ylab='')
  lines(c(-5,5),c(0,0),lty=2)
  lines(c(0,0),c(-5,5),lty=2)
  arrows(0,0, s[i]*v1[1],s[i]*v1[2],pch=19,lwd=2,col='blue')
  if(i==length(s)){text(2*v1[1],2*v1[2],expression(2*v[1]),pos=4,cex=2,col='blue')}
  arrows(0,0, v1[1],v1[2],pch=19,lwd=2)
  text(v1[1],v1[2],expression(v[1]),pos=4,cex=2)
  
}
```

---
  
#Scalar multiply by 0.5
  
```{r smultgeom2, fig.align='center',echo=FALSE,fig.show='animate',interval=0.5,aniopts='controls', fig.cap='Click to see',ffmpeg.format='ogg'}
v1<-c(1,2)
s<-seq(1,0.5,-0.1)
for (i in 1:length(s)){
  plot(0,0,'n',xlim=c(-5,5),ylim=c(-5,5),xlab='',ylab='')
  lines(c(-5,5),c(0,0),lty=2)
  lines(c(0,0),c(-5,5),lty=2)
  arrows(0,0, v1[1],v1[2],pch=19,lwd=2)
  text(v1[1],v1[2],expression(v[1]),pos=4,cex=2)
  arrows(0,0, s[i]*v1[1],s[i]*v1[2],pch=19,lwd=2,col='blue')
  if(i==length(s)){text(0.5*v1[1],0.5*v1[2],expression(0.5*v[1]),pos=4,cex=2,col='blue')}
}
```


---

#Scalar multiplication by 2

```{r multpts,  fig.align='center',echo=FALSE,fig.show='animate',interval=0.5, fig.cap='Click to see',ffmpeg.format='ogg'}
grid2<-rbind(rnorm(50),rnorm(50))
s<-seq(1,2,by=0.2)
for (j in 1:length(s)){
  plot(0,0,'n',xlim=c(-5.1,5.1),ylim=c(-5.1,5.1),xlab='',ylab='')
  lines(c(-5.1,5.1),c(0,0),lty=2)
  lines(c(0,0),c(-5.1,5.1),lty=2)
  vsv<-s[j]*grid2
  points(vsv[1,],vsv[2,],pch=19,lwd=2,col='blue')
}
```

---

#Scalar multiplication by 0.5

```{r multpts2,  fig.align='center',echo=FALSE,fig.show='animate',interval=0.5,fig.cap='Click to see',ffmpeg.format='ogg'}
grid2<-rbind(rnorm(50),rnorm(50))
s<-seq(1,0.5,by=-0.1)
for (j in 1:length(s)){
  plot(0,0,'n',xlim=c(-5.1,5.1),ylim=c(-5.1,5.1),xlab='',ylab='')
  lines(c(-5.1,5.1),c(0,0),lty=2)
  lines(c(0,0),c(-5.1,5.1),lty=2)
  vsv<-s[j]*grid2
  points(vsv[1,],vsv[2,],pch=19,lwd=2,col='blue')
}
```

---
  
# Scalar multiplication by 2
  
$$\begin{aligned}{\color{red}{2}}{\mathbf{v}}_1&={\color{red}{2}}\begin{bmatrix}1\\2\end{bmatrix}\\&=\begin{bmatrix}{\color{red}{2}}\times 1\\{\color{red}{2}}\times 2\end{bmatrix}\\&=\begin{bmatrix}2\\4\end{bmatrix}\end{aligned}$$
  
---
  
# Scalar multiplication by 0.5
  
$$\begin{aligned}{\color{red}{0.5}}{\mathbf{v}}_2&={\color{red}{0.5}}\begin{bmatrix}3\\1\end{bmatrix}\\&=\begin{bmatrix}{\color{red}{0.5}}\times 3\\{\color{red}{0.5}}\times 1\end{bmatrix}\\&=\begin{bmatrix}1.5\\0.5\end{bmatrix}\end{aligned}$$
  
---
  
# Scalar multiplication and data
  
- How might scalar multiplication be relevant to data?<!--D-->
--

- Suppose trade deficit is measured in Chinese Yuan instead of $US where 6 
Yuan = $US1.<!--D-->
--

- Then we can get the trade defcit in Yuan by multiplying by the scalar 6.<!--D-->
--

- Suppose Australia had a trade deficit of $US1bn in 2016 
and $US3bn in 2017.<!--D-->
--

  + This is equivalent to a trade deficit of 6bn Yuan in 2016 and 18bn Yuan in 2017.

---
  
# Basis vector
  
-  Every vector can be created using<!--D-->
--

  + Scalar multiplication 
  + Vector addition 
  + Two *basis* vectors<!--D-->
--

- The basis vectors we use are:
  
$${\mathbf{b}}_1=\begin{bmatrix}1\\0\end{bmatrix}\quad\mbox{and}\quad{\mathbf{b}}_2=\begin{bmatrix}0\\1\end{bmatrix}$$
  
---
  
# How the basis works
  
- Suppose we want the vector $\begin{bmatrix}x\\y\end{bmatrix}$
- This is given by

$$x\mathbf{b}_1+y\mathbf{b}_2$$
  or
$$x\begin{bmatrix}1\\0\end{bmatrix}+y\begin{bmatrix}0\\1\end{bmatrix}$$
  
---
  
# Build a vector
  
```{r basis,  fig.align='center',echo=FALSE,fig.show='animate',interval=0.5,aniopts='controls', fig.cap='Click to see',ffmpeg.format='ogg'}

sx<-seq(1,2,by=0.2)
for (j in 1:length(sx)){
  plot(0,0,'n',xlim=c(-3.1,3.1),ylim=c(-3.1,3.1),xlab='',ylab='',asp=1)
  lines(c(-3.1,3.1),c(0,0),lty=2)
  lines(c(0,0),c(-3.1,3.1),lty=2)
  arrows(0,0,sx[j],0,pch=19,col = 'red',lwd=2)
  if(j==length(sx)){text(sx[j],0,expression(2*b[1]),col='red',pos=4,cex=2)}
  arrows(0,0,1,0,pch=19,col = 'blue',lwd=2)
  text(1,0,expression(b[1]),col='blue',pos=4,cex=2)
  arrows(0,0,0,1,pch=19,col = 'blue',lwd=2)
  text(0,1,expression(b[2]),col='blue',pos=4,cex=2)
}
sy<-seq(1,3,by=0.4)
for (j in 1:length(sy)){
  plot(0,0,'n',xlim=c(-3.1,3.1),ylim=c(-3.1,3.1),xlab='',ylab='',asp=1)
  lines(c(-3.1,3.1),c(0,0),lty=2)
  lines(c(0,0),c(-3.1,3.1),lty=2)
  arrows(0,0,2,0,pch=19,col = 'red',lwd=2)
  text(2,0,expression(2*b[1]),col='red',pos=4,cex=2)
  arrows(0,0,0,sy[j],pch=19,col = 'red',lwd=2)
  if(j==length(s)){text(0,sy[j],expression(3*b[2]),col='red',pos=4,cex=2)}
  arrows(0,0,0,1,pch=19,col = 'blue',lwd=2)
  text(0,1,expression(b[2]),col='blue',pos=4,cex=2)
}
sadd<-seq(0,2,by=0.4)
for (j in 1:length(sy)){
  plot(0,0,'n',xlim=c(-3.1,3.1),ylim=c(-3.1,3.1),xlab='',ylab='',asp=1)
  lines(c(-3.1,3.1),c(0,0),lty=2)
  lines(c(0,0),c(-3.1,3.1),lty=2)
  arrows(0,0,2,0,pch=19,col = 'red',lwd=2)
  text(2,0,expression(2*b[1]),col='red',pos=4,cex=2)
  arrows(0,0,0,3,pch=19,col = 'red',lwd=2)
  text(0,3,expression(3*b[2]),col='red',pos=4,cex=2)
  arrows(sadd[j],0,sadd[j],3,pch=19,col = 'black',lwd=2)
}
plot(0,0,'n',xlim=c(-3.1,3.1),ylim=c(-3.1,3.1),xlab='',ylab='',asp=1)
lines(c(-3.1,3.1),c(0,0),lty=2)
lines(c(0,0),c(-3.1,3.1),lty=2)
arrows(0,0,2,0,pch=19,col = 'red',lwd=2)
text(2,0,expression(2*b[1]),col='red',pos=4,cex=2)
arrows(0,0,0,3,pch=19,col = 'red',lwd=2)
text(0,3,expression(3*b[2]),col='red',pos=4,cex=2)
arrows(0,0,2,3,pch=19,col = 'black',lwd=2)
text(2,3,'(2,3)',col='black',pos=4,cex=2)
```

---
class: inverse, center, middle

# Matrices

---
  
# Matrices
  
- You have learnt about matrices before as an array of numbers.<!--D-->
--

- Today think about matrices as *transforming* points in space.<!--D-->
--

- By multiplying a vector by a matrix, points in space are stetched, rotated and flipped around<!--D-->
--

- But what does this have to do with an array of numbers?
  
---
  
# Example of a matrix
  
- Consider the following matrix
$$\mathbf{M}=\begin{bmatrix}5 &1\\2 &4\end{bmatrix}$$
- Think of this as something that moves all the points in space to a new address.
- It encodes where $\mathbf{b}_1$ and $\mathbf{b}_2$ get moved to.

---
  
# Change of basis 
  
Multiplying by the matrix $\mathbf{M}=\begin{bmatrix}{\color{magenta} 5} &\color{green}{1}\\{\color{magenta} 2} &{\color{green}4}\end{bmatrix}$ will: 
  
Move the vector $\mathbf{b}_1=\begin{bmatrix}1\\0\end{bmatrix}$ to $\mathbf{b}^{*}_1=\begin{bmatrix}{\color{magenta} 5}\\{\color{magenta} 2}\end{bmatrix}$
  
Move the vector $\mathbf{b}_2=\begin{bmatrix}0\\1\end{bmatrix}$ to $\mathbf{b}^{*}_2=\begin{bmatrix}{\color{green}  1}\\{\color{green} 4}\end{bmatrix}$
  
Matrix multiplication is a very different type of multiplication than scalar multiplication.

---
  
# Change of basis 
  
```{r basischange,  fig.align='center',echo=FALSE}
plot(0,0,'n',xlim=c(-5.1,5.1),ylim=c(-5.1,5.1),xlab='',ylab='',asp=1)
lines(c(-5.1,5.1),c(0,0),lty=2)
lines(c(0,0),c(-5.1,5.1),lty=2)
arrows(0,0,5,2,pch=19,col = 'red',lwd=2)
text(5,2,expression(b[1]^'*'),col='red',pos=4,cex=2)
arrows(0,0,1,4,pch=19,col = 'red',lwd=2)
text(1,4,expression(b[2]^'*'),col='red',pos=4,cex=2)
arrows(0,0,1,0,pch=19,col = 'blue',lwd=2)
text(1,0,expression(b[1]),col='blue',pos=4,cex=2)
arrows(0,0,0,1,pch=19,col = 'blue',lwd=2)
text(0,1,expression(b[2]),col='blue',pos=4,cex=2)
```

---
  
#Old Basis
  
$$\begin{aligned}
\mathbf{v}&=x\mathbf{b}_1+y\mathbf{b}_2\\
&=x\begin{bmatrix}1\\0\end{bmatrix}+y\begin{bmatrix}0\\1\end{bmatrix}\\
&=\begin{bmatrix}x\\y\end{bmatrix}
\end{aligned}$$
  
---
  
# New Basis
  
$$\begin{aligned}
\mathbf{v}^*&=x\mathbf{b}^{*}_1+y\mathbf{b}^{*}_2\\
&=x\begin{bmatrix}5\\2\end{bmatrix}+y\begin{bmatrix}1\\4\end{bmatrix}\\
&=\begin{bmatrix}5x+y\\2x+4y\end{bmatrix}
\end{aligned}$$
  
---
  
# In general
  
Multiplying by the matrix $\mathbf{M}$ moves all vectors $\mathbf{v}$ to $\mathbf{v}^*$
  
We can write this as:
  
$$\mathbf{v}^*=\mathbf{Mv}=\begin{bmatrix}5 &1\\2 &4\end{bmatrix}\begin{bmatrix}x\\y\end{bmatrix}=\begin{bmatrix}5x+y\\2x+4y\end{bmatrix}$$
  
Another way to think of it is as multiplying the rows of ${\mathbf M}$ by the column $\mathbf {v}$


---
  
# Matrix Multiplication in General

```{r matmult,  fig.align='center',echo=FALSE,fig.show='animate',interval=0.5,aniopts='controls',fig.cap='Click to see',ffmpeg.format='ogg'}
g=10
grid2<-rbind(rnorm(80),rnorm(80))
cols<-rep('black',80)
cols[grid2[1,]*grid2[2,]>0]<-'green4'
m11<-1
m12<-0.2
m21<-0.2
m22<-1
M<-matrix(c(m11,m21,m12,m22),2,2)
s<-svd(M)
thU<-asin(s$u[2,1])
thV<-asin(s$v[2,1])
td1<-s$d[1]
td2<-s$d[2]
incU<-seq(0,thU,length.out = g)
incV<-seq(0,thV,length.out = g)
inctd1<-seq(1,td1,length.out = g)
inctd2<-seq(1,td2,length.out = g)
bb<-10

for (j in 1:g){
  plot(0,0,'n',xlim=c(-3.1,3.1),ylim=c(-3.1,3.1),xlab='',ylab='',asp=1)
  lines(c(-5.1,5.1),c(0,0),lty=2)
  lines(c(0,0),c(-5.1,5.1),lty=2)  
  Ui<-matrix(c(-cos(incU[j]),sin(incU[j]),sin(incU[j]),cos(incU[j])),2,2)
  Vi<-matrix(c(-cos(incV[j]),sin(incV[j]),sin(incV[j]),cos(incV[j])),2,2)
  Di<-matrix(c(inctd1[j],0,0,inctd2[j]),2,2)
  Mi<-Ui%*%Di%*%t(Vi)
  gridi<-Mi%*%grid2
  points(gridi[1,],gridi[2,],pch=19,cex=2,col=cols)
  arrows(0,0,m11,m21,pch=19,col = 'red',lwd=3)
  lines(c(bb*m11,-bb*m11),c(bb*m21,-bb*m21),lty=2,col='red')
  text(m11,m21,expression(b[1]^'*'),col='red',pos=4,cex=2)
  arrows(0,0,m12,m22,pch=19,col = 'red',lwd=3)
  lines(c(bb*m12,-bb*m12),c(bb*m22,-bb*m22),lty=2,col='red')
  text(m12,m22,expression(b[2]^'*'),col='red',pos=4,cex=2)
  arrows(0,0,1,0,pch=19,col = 'blue',lwd=3)
  text(1,0,expression(b[1]),col='blue',pos=4,cex=2)
  arrows(0,0,0,1,pch=19,col = 'blue',lwd=3)
  text(0,1,expression(b[2]),col='blue',pos=4,cex=2)
  
}
```

---

  
# Matrix Multiplication and Data
  
An simple application of matrix multiplication is to create *new* variables.

- Suppose we want a matrix to transform :
  + The number of chickens and number of rabbits to...<!--D-->
--

  + The number of heads and the number of feet.

---

# Answer

- First basis vector is the number of heads and feet of each chicken.
- Second basis vector is the number of heads and feet of each rabbit.

$$\begin{bmatrix}1&1\\2&4\end{bmatrix}$$
  
---

# Another one
  
- Suppose we want a matrix to transform:
  + The trade deficit in 2016 and trade deficit in 2017 to...<!--D-->
--

  + The average trade deficit from 2016 to 2017 and the change in the trade deficit between 2016 and 2017<!--D-->
--


What does this matrix look like?  Try yourselves.

---

# Answer

- First basis vector average and change if trade deficit was $1 in 2016 
and $0 in 2017.
- Second basis vector average and change if trade deficit was $0 in 2016 
and $1 in 2017.

$$\begin{bmatrix}0.5&0.5\\-1&1\end{bmatrix}$$  

---


# Non-Square Matrices
  
- So far we have only considered $2\times 2$ matrices.
- These move every point in 2D space to another point in 2D space
- However we can have a matrix with 3 rows and 2 columns (or $3\times2$ matrix).

$$\mathbf{M}=\begin{bmatrix}1 &2\\3 &4\\5 &6\\\end{bmatrix}$$
  
---
  
# Non-square Matrices
  
$$\begin{aligned}\mathbf{Mz}&=\begin{bmatrix}1 &2\\3 &4\\5 &6\\\end{bmatrix}\begin{bmatrix}x\\y\end{bmatrix}\\&=x\begin{bmatrix}1\\3\\5\end{bmatrix}+y\begin{bmatrix}2\\4\\6\end{bmatrix}\\&=\begin{bmatrix}x+2y\\3x+4y\\5x+6y\\\end{bmatrix}\end{aligned}$$
  
---
  
# Non-square matrices
  
- Matrix multiplication works in the same way<!--D-->
--

- However the *input* is a 2-dimensional vector but the *output* is a 3-dimensional vector<!--D-->
--

- Multiplying by a non-square matrix changes the dimension
--

- The points will still lie in a two-dimensional *subspace* of three-dimensional space.
---

# In factor analysis

- For an example of this consider the factor model and in particular the component $\boldsymbol{\Lambda}\boldsymbol{f}_i$.
--

- This takes a low-dimensional set of factors and maps them to a high-dimensional vector of data.
--

- What about multiplying by a $2\times 3$ matrix?

---
  
  
# Dimension reduction

- Multiplying by a $2\times 3$ matrix takes a $3$-dimensional vector and maps it to a point in $2$-dimensional space.
--

- This is an example of *dimension reduction*
--

- For instance taking the first two principal component of data is the same as multiplying the data by a $2\times p$ matrix

---

# Matrix composition
  
- What does it mean to multiply matrices?  
--

- What does $\mathbf{F}\mathbf{G}$ when $\mathbf{G}$ is a $4\times 2$ matrix and $\mathbf{F}$ a $3\times 4$ matrix?
--

- It is as a *composition* of transformations:<!--D-->
--

  + Transform a $2$-dimensional vector $\mathbf{v}$ into a $4$-dimensional vector by multiplying by the $4\times 2$ matrix $\mathbf{G}$<!--D-->
--

  + Transorm the result into a $3$-dimensional vector by multiplying by a $3\times 4$ matrix $\mathbf{F}$
  
---
  
# Mathematically
  
- When reading below go from right to left:
  
$$\underset{{ (3\times 2)}}{\mathbf{H}}
\underset{{ \leftarrow}}{\times}
\underset{{ (2\times 1)}}{\mathbf{v}}
\underset{=}{{ =}}
\underset{{ (3\times 4)}}{\mathbf{F}}
\underset{{ \leftarrow}}{\times}
\underset{{ (4\times 2)}}{\mathbf{G}}
\underset{{ \leftarrow}}{\times}
\underset{{ (2\times 1)}}{\mathbf{v}}$$
  
- The matrix $\mathbf{H}=\mathbf{FG}$ is a composition of doing $\mathbf{G}$ and then $\mathbf{F}$ 
  
---
  
# Matrix multiplication
  
- What happens if you try to multiply $\mathbf{GFv}$?<!--D-->
--

$$\underset{{ (4\times 2)}}{\mathbf{G}}\underset{{ \not\leftarrow}}{\times}\underset{{ (3\times 4)}}{\mathbf{F}}\underset{{ \not\leftarrow}}{\times}\underset{{ (2\times 1)}}{\mathbf{v}}$$<!--D-->
--

- Going from right to left, $\mathbf{F}$ will try to transform a 4-dimensional vector to a 3-dimensional vector and cannot take the is a 2-dimensional vector $\mathbf{v}$  as its input!<!--D-->
--

- Also $\mathbf{G}$ transforms a 2-dimensional vector to a 4-dimensional vector so cannot take the 3-dimensional output of $\mathbf{F}$ as its input.

---
  
# Matrix Multiplication
  
- Although it is possible to get $\mathbf{FG}$ it is impossible to get $\mathbf{GF}$.<!--D-->
--

- In general, for matrix multiplication $\mathbf{AB}\neq\mathbf{BA}$ even for square matrices.<!--D-->
--

- Matrix multiplication is *non-commutative*.

---
class: inverse, center, middle

# Inner products and rotations

---

# Inner product

- One final operation with vectors is called the *inner product*.
--

- The inner product of two vectors $\mathbf{x}$ and $\mathbf{y}$ is defined as $\mathbf{x}'\mathbf{y}$.
--

- The inner product of a vector with itself $\mathbf{x}'\mathbf{x}$ is the length of the vector squared. 
  
---
# Orthogonality

- An interesting case is when the inner product between two vectors is zero.
  - $\mathbf{x}'\mathbf{y}=0$
- In this case, the vectors $\mathbf{x}$ and $\mathbf{y}$ are at a 90 degree angle to one another.
- This is also called *orthogonality*.

---
# Rotation matrix
  
- Another important matrix is known as a *orthogonal rotation* matrix.<!--D-->
--

- It spins all the points around keeping them the same distance from the origin and from each other.<!--D-->
--

- The word *orthogonal* means that the new basis vectors are at right angles to one another.

---
  
# Rotation Matrix

- In matrix algebra rotation can be defined in any dimension.
- It is easiest to understand in two dimensions.
- An example of a rotation matrix in two dimensions is:
$$\mathbf{R}=\begin{bmatrix}0.8 & -0.6\\0.6 & 0.8\end{bmatrix}$$


---

#Rotation Matrix

```{r matmultr,  fig.align='center',echo=FALSE,fig.show='animate',interval=0.5,aniopts='controls', fig.cap='Click to see',ffmpeg.format='ogg'}
g=10
grid2<-rbind(rnorm(80),rnorm(80))
cols<-rep('black',80)
cols[grid2[1,]*grid2[2,]>0]<-'green4'
m11<-0.8
m12<--0.6
m21<-0.6
m22<-0.8
M<-matrix(c(m11,m21,m12,m22),2,2)
s<-svd(M)
thU<-asin(s$u[2,1])
thV<-asin(s$v[2,1])
td1<-s$d[1]
td2<-s$d[2]
incU<-seq(0,thU,length.out = g)
incV<-seq(0,thV,length.out = g)
inctd1<-seq(1,td1,length.out = g)
inctd2<-seq(1,td2,length.out = g)
bb<-10

for (j in 1:g){
  plot(0,0,'n',xlim=c(-3.1,3.1),ylim=c(-3.1,3.1),xlab='',ylab='',asp=1)
  lines(c(-5.1,5.1),c(0,0),lty=2)
  lines(c(0,0),c(-5.1,5.1),lty=2)  
  Ui<-matrix(c(-cos(incU[j]),sin(incU[j]),sin(incU[j]),cos(incU[j])),2,2)
  Vi<-matrix(c(-cos(incV[j]),sin(incV[j]),sin(incV[j]),cos(incV[j])),2,2)
  Di<-matrix(c(inctd1[j],0,0,inctd2[j]),2,2)
  Mi<-Ui%*%Di%*%t(Vi)
  gridi<-Mi%*%grid2
  points(gridi[1,],gridi[2,],pch=19,cex=1.5,col=cols)
  arrows(0,0,m11,m21,pch=19,col = 'red',lwd=3)
  lines(c(bb*m11,-bb*m11),c(bb*m21,-bb*m21),lty=2,col='red')
  text(m11,m21,expression(b[1]^'*'),col='red',pos=4,cex=2)
  arrows(0,0,m12,m22,pch=19,col = 'red',lwd=3)
  lines(c(bb*m12,-bb*m12),c(bb*m22,-bb*m22),lty=2,col='red')
  text(m12,m22,expression(b[2]^'*'),col='red',pos=4,cex=2)
  arrows(0,0,1,0,pch=19,col = 'blue',lwd=3)
  text(1,0,expression(b[1]),col='blue',pos=4,cex=2)
  arrows(0,0,0,1,pch=19,col = 'blue',lwd=3)
  text(0,1,expression(b[2]),col='blue',pos=4,cex=2)
}
```

---
  
# Rotation matrix
  
- The new basis vectors should have a length of 1 therefore $\mathbf{r}_i'\mathbf{r}_i=1$ where $\mathbf{r}_i$ are the columns of $\mathbf{R}$.
--

- The new basis vectors should be at right angles $\mathbf{r}_i'\mathbf{r}_j=0$
--

- A rotation matrix $\mathbf{R}$ has the property $\mathbf{R'R}=\mathbf{I}$<!--D-->

---

# Rotations in PCA

- In PCA the weights of the principal components have two properties
  - $\mathbf{w}_i'\mathbf{w}_i=1$
  - $\mathbf{w}_i'\mathbf{w}_j=0$
- All $p$ principal components can be obtained by multiplying by a rotation matrix by $p$-dimensional data.
- Taking principal components *rotates* the data.

---

# Rotations in Factor Analysis

- In factor analysis rotation was used to obtain more interpretable factors.
- This was a $r$-dimensional rotation of the factors rather than the data.
- Again this involved multiplying by a rotation matrix.

---

# Conclusion

- Understand the geometry of matrix algebra assists in understanding PCA and Factor analysis in particular.
- PCA forms new variables as linear combinations of old variables.
  - This corresponds to matrix multiplication.
- The role of *rotation* should now also be understood.
- The next step is to understand the importance of matrix decompositions in the methods covered so far.