---
title: "Getting Started with R"
subtitle: "High Dimensional Data Analysis"
author: "Anastasios Panagiotelis"
date: "Lecture 2"
output:
  xaringan::moon_reader:
    chakra: libs/remark-latest.min.js
    lib_dir: libs
    css: [default,"mtheme.css","mod.css"]
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false

---

class: inverse, center, middle

# Basics

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = FALSE, fig.align = "center")
require(magrittr)
require(tidyverse)
require(plotly)
require(widgetframe)
require(animation)
require(scatterplot3d)
require(DT)
require(rgl)
require(knitr)
require(kableExtra)
rm(list=ls())
```

---


# The R project

- In this unit we use the software package R.
--

- R is one of the the most popular software tool for professionals in the fields of Business Analytics/Data Science.

---

# History of R

- R is based on earlier statistical software called 'S' which was developed at Bell Labs in the 1970s.
--

- R was initially developed in the early 1990s by two academics at the University of Auckland, Ross Ihaka and Robert Gentlemen. 
--

- R has grown substantially since then and is now supported the not-for-profit R Foundation

---

# R is Free Software

- R is free in two ways
  - It doesn't cost any money.
  - All of the source code of R is available meaning R can be customised, modified, and most importantly extended.
- R is part of a big Free Software project known as the GNU Project

---

# Downloading R and R Studio

- R can be downloaded [here](http://cloud.r-project.org/)
--

- Exact details of installing R will depend on whether you use Windows, Mac or Linux.
--

- A great tool for both new and experienced users of R is *RStudio*.
--

- It can be downloaded [here](https://www.rstudio.com/products/RStudio/) free of cost.
--

- If you haven't done so already try to download R and R Studio.

---

# Ways to use R

- To keep track of your workflow use a **script**:
  - Open a new script by typing Ctrl+Shift+N
  - Run a single line of code by pressing Ctrl+Enter
  - Run a whole script by pressing Ctrl+Shift+S or Ctrl+Shift+Enter
- You can save scripts to run them anytime.
--

- Scripts allow you to keep analysis **replicable** which is important in research and business.

---

# Variables in R

- In R everything is stored in a **variable**. 
--

- Here the word variable has a slightly different meaning to the usual statistical meaning. 
--

- In R, think of these as little boxes with names on them.
--

- We can put a number into these boxes, or words or matrices or entire blocks of data or even other boxes.

---

# Assigning Variables

- How to store the number 1 in a variable `a` and the number 2 in a variable `b`?
```{r, echo=TRUE}
a<-1
b=2
```
- Note that you can use either `<-` or `=` to assign variables. 

---

# Seeing results

To see what is stored in a variable

```{r, echo=TRUE}
print(b)
str(b)
```

---

# Character variables

- Text can also be stored in a variable.  
- Try to store your own name in a variable called *name*.
```{r, echo=TRUE,eval=TRUE}
name<-'Anastasios'
str(name)
```
- Use apostrophes so that R does not look for a variable called `Anastasios`.

---

# Workspace

- All variables are kept in the **workspace**.  You can see what is in your workspace by using the command

```{r,echo=TRUE}
ls()
```


---

# Clear Workspace

- You can clear the workspace using

```{r,eval=FALSE,echo=TRUE}
rm(list=ls())
```
- If you try ls() again the workspace will be empty.  In RStudio you  can also see all the variables in the *Environment* tab.
--

- It is good practice to start every script with this command so that you do not accidentally use data from a different project.

---

# Working directory

- If you read data stored on your computer, or if you save plots or data then the concept of a **working directory** is important.  
--

- To check your working directory type

```{r,eval=FALSE,echo=TRUE}
getwd()
```

- To change the Working directory use `setwd` 

```{r,eval=FALSE,echo=TRUE}
setwd("/home/anastasios/Documents")
```

---

# Basic arithmetic in R

- Basic arithmetic is fairly simple.  Try `a+b`.  Also we will put this in a new variable called `z`.

```{r,echo=TRUE}
z<-a+b
str(z)
```

- To subtract use `-`, to multiply use `*`, to divide `/` and to take powers use `^`.

---

# Functions in R

- Apart from very simple arithmetic, variables in R are manipulated using a **function**.  
--

- The input goes in parentheses, while the output is assigned to a new variable.  
--

- For example the function *sqrt* is used to take the square root.  

```{r, eval=TRUE, echo=TRUE}
rootb<-sqrt(b)
str(rootb)
```

---

#Garbage in/Garbage out

What happens when you take a square root of something that is not a number?

```{r, error=TRUE, eval=TRUE,echo=TRUE}
rootname<-sqrt(name)
```

Many if not most of the mistakes you make in R occur because you enter the incorrect type of input in a function.
---

# Getting Help

- If you aren't sure what a function does, use R help.  The easiest way is to simply use the `?`

```{r, echo=TRUE,eval=FALSE}
?sqrt
```

- If you want to do something and do not know the name of the relevant function you can search using `??`.  
--

- Find a function to do logarithms using

```{r, echo=TRUE,eval=FALSE}
??logarithms
```

---

# Comments

Anything after a `#` will not be executed by R.
```{r, error=TRUE, eval=TRUE, echo=TRUE}
a<-1 # Set the variable a to 1
#x<-4 This line is not executed 
str(a)
str(x)
```
Comment multiple lines using Ctrl+Shift+C

---

# Vectors

In stats we have many observations for each variable.  The function `c()` stores these in a **vector**. Suppose we have the following data:

Names  | Drink | Consumption | Satisfaction
------------- | ------------- | ------------- | ------------- 
Andrew  | Coke | 50 | 5
Boris  | Pepsi | 40 | 4
Cathy  | Coke | 25 | 4
Diana  | 7Up | 0 | 3


---

# Manually inputting data

First let's create the variable Consumption
```{r, echo=TRUE}
Consumption<-c(50,40,25,0)
print(Consumption)
str(Consumption)
```

Put values of drink into a variable.

---

# Solution

The solution is
```{r, echo=TRUE, eval=TRUE}
Drink<-c('Coke','Pepsi','Coke','7Up')
print(Drink)
str(Drink)
```

---

# Vector

These variables are example of a **vector**.  Sometimes when we apply a function to a vector, we apply the function to each element.

```{r, echo=TRUE}
logcons<-log(Consumption)
str(logcons)
```


---

# Vectors

Other functions take a vector as an input and return a single number as the output

```{r, echo=TRUE}
meancons<-mean(Consumption)
str(meancons)
```

---

# Inf and NaN

The values `Inf` and `-Inf` refer to positive and negative infinity.  The value `NaN` stands for not a number and indicates an error.

```{r, echo=TRUE}
log(-1)
```

It is important to distinguish `NaN` from `NA`. The latter is used for missing data.

---

# Data Frame

- It is tedious to manually enter large datasets in this way.  You will usually import data from an external file.
--

- There are many ways to import data.  For files with the `.rds` extension it is easy

```{r, echo=TRUE}
Beer<-readRDS("Beer.rds")
```

- Get the location of the file right.  You can also open a file through the *file* tab

---

# Data Frame

- There is only one variable here, the variable *Beer*.  However this is a very special case of variable known as a **Data Frame**.
--

- A data frame contains other variables.  For example *alcohol* content can be accessed via.

```{r, echo=TRUE}
str(Beer$alcohol)
```

---

# Lists

Another object common in R is known as a **list**.  A list can contain completely different variables.
```{r, echo=TRUE}
alist<-list(w=name, x=Drink, y=Beer)
```
elements of lists are accessed using `[[]]` or `$`
```{r, echo=TRUE}
alist[[1]]
```

---

class: center, inverse middle

# Packages

---

# R Packages

- One of the biggest advantages of R is the use of add-on packages, which are are easily downloaded from an online repository called **CRAN**.  Using a package involves two steps:
--

  - Downloading and installing the package using the function `install.package`.
  - Load the package using `library` function
--

- Both these steps can also be done in RStudio through the *Packages* tab.
--

- Try install and load the R package `ggplot2`

---

# Package Documentation

- By downloading the package you also download all of the help documentation.

```{r, eval=FALSE,echo=TRUE}
install.packages('ggplot2')
```
To load the package
```{r,echo=TRUE}
library(ggplot2)
```

---

# Graphics in R

- Three different ways to do graphs in R
  - Base graphics do not require any packages
  - Trellis graphics (using package lattice)
  - ggplot2
--

- In this unit you will mostly be given instruction on using `ggplot2`, however if you have learnt base graphics in another unit and prefer this, then you can use it.
--

- There are many resources for learning `ggplot2`, including some that are free online.

---

# MT cars dataset

To demonstrate `ggplot2` we use the dataset `mpg`, which contains information on the fuel efficiency of different cars.  This can be loaded into R using the command

```{r,echo=TRUE}
data(mpg)
```

It is data that comes together with `ggplot2`.

---

# ggplot object

- To make a plot, the first task is to create a ggplot object.  
--

- You need to specify the data frame and the **aesthetic**.  
--

- In a 2D plot we can compare two variables
--

- To start, think of the aesthetic as the x and y variable.
--

- Consider a plot to compare the fuel efficiency of a car in the city and on the highway.

---

# Number of Cylinders v MPG

```{r,fig.align='center',echo=TRUE,fig.height=6}
ggplot(mpg,aes(x=cty,y=hwy))
```

---

# Geometry

- This should produce some axes and labels but there is no plot yet.
--

- To produce a plot we need to tell R what type of plot we want.
--

- In the language of ggplot this is called a **geometry**
--

- For a scatter plot the geometry is `geom_point()`.

---

# Scatterplot in R

```{r,fig.align='center',echo=TRUE,fig.height=6}
ggplot(mpg,aes(x=cty,y=hwy))+geom_point()
```

---

# Other aesthetics

We can think of colour as a third aesthetic

```{r,fig.height=4,fig.align='center',echo=TRUE}
ggplot(mpg,aes(x=cty,y=hwy,col=cyl))+geom_point()
```

---

# With a categorical variable

```{r,fig.align='center',echo=TRUE, fig.height=6}
ggplot(mpg,aes(x=cty,y=hwy,col=drv))+geom_point()
```

---

# Quicker plots

A useful function in `ggplot2` is to use `qplot` which will try to guess the plot you want.  Try these examples  

```{r,eval=FALSE,fig.align='center',echo=TRUE}
qplot(x=cty,y=hwy,data=mpg)
qplot(x=cty,data=mpg)
```

---

# Exporting graphics

- You can export graphics using the *Export* tab in the *Plot* tab in Rstudio.  
--

- Many different file formats are available. As an alternative you can do the following:

```{r,eval=FALSE,echo=TRUE}
pdf('myplot.pdf')
qplot(x=cty,data=mpg)
dev.off()
```

- Other file formats such an png or jpeg can be used instead of pdf.

---

# Data Manipulation

- There are several ways to manipulate data, but a particularly useful and easy package to use is called **dplyr**.
--

- We can exclude observations using the `filter` function.
--

- To really understand how to use this function it helps to know about **logical operators** (try `?Logic`) and **relational operators** (try `?Comparison`).
--

- We will do a few simple examples here

---

# Using dplyr

To create a new data frame that only includes 4 wheel drives

```{r,message=FALSE,echo=TRUE}
library(dplyr)
mpg_4wd<-filter(mpg,drv=='4')
```

To exclude all 4 wheel drives

```{r,echo=TRUE}
mpg_no4wd<-filter(mpg,drv!='4')
```

---

# Two conditions

Suppose we only want to consider cars that are 4 wheel drives **and** can drive more than 15 miles per gallon on the highway

```{r,echo=TRUE}
mpg_4wd_hwyg15<-filter(mpg,(drv=='4')&(hwy>15))
```

Or those that are either 4 wheel drives **or** can drive less than 15 miles per gallon in the city

```{r,echo=TRUE}
mpg_4wd_ctyl15<-filter(mpg,(drv=='4')|(cty<15))
```

---

# Without dplyr

- This sort of data manipulation can be done without `dplyr` but is more verbose.  
- For example the last line would be
```{r,echo=TRUE}
mpg_4wd_ctyl15<-mpg[((mpg$drv=='4')|(mpg$cty<15)),]
```
- Both give the same result, use whichever you prefer.

---

# Summarise Fuction

- Suppose we want the mean and standard deviation of the (filtered) data.

```{r,echo=TRUE}
mpg_4wd_ctyl15<-filter(mpg,(drv=='4')|(cty<15))
mean_sd_hwy<-summarise(mpg_4wd_ctyl15,mean(hwy),sd(hwy))
mean_sd_hwy
```


---

# Pipes

- Pipes from the *magrittr* package make this easier.

```{r,echo=TRUE}
filter(mpg,(drv=='4')|(cty<15))%>%
  summarise(mean(hwy),sd(hwy))%>%
  print
```


---

# Conclusion

- This lecture has given you a foundation in R
--

- You can use R to do much more including collecting data off the web, cleaning the data, fitting models to the data, creating web applications, or even creating documents (these slides were created in RStudio).
--

- This can be daunting, but remember the best thing about R is that there are lots of ways to teach yourself R.
