---
title: "Introduction and Motivation"
subtitle: "High Dimensional Data Analysis"
author: "Anastasios Panagiotelis"
date: "Lecture 1"
output:
  xaringan::moon_reader:
    chakra: libs/remark-latest.min.js
    lib_dir: libs
    css: [default,"mtheme.css","mod.css"]
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false

---

class: inverse, center, middle

# Housekeeping

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = FALSE, fig.align = "center")
require(magrittr)
require(tidyverse)
require(plotly)
require(widgetframe)
require(animation)
require(scatterplot3d)
require(DT)
require(rgl)
require(knitr)
require(kableExtra)
require(broom)
require(ca)
require(ggrepel)

```

---


# Welcome to HDDA!

- Lecturer for the unit
  - Anastasios Panagiotelis
- Use the moodle forum to ask questions
    + Feel free to answer each others questions.
    + I will also provide answers.
- Email anastasios.panagiotelis@monash.edu only for issues that are personal.
- For details on consultation see Moodle  

---

# High-Dimensional Data?

- First what do we mean by *High Dimensional*?
- The data we look at will have:
  + Observations
  + Variables
- Generally *High Dimensional* implies that the number of *variables* is large.
--

- The term, high-dimensional also relates thinking about and visualising data as points in space.

---
class: inverse, middle, center

# A Data Story

---

# US States

- Five indicators of the quality of life in the 50 States of the USA in 1977.
  - Income,
  - Illiteracy rate,  
  - High school graduation rate,
  - Life expectancy,
  - Murder rates.
- Let's explore!
---

# A dataset

```{r, echo=FALSE}
States<-readRDS('StateSE.rds')
kable(States)%>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"))%>%
  scroll_box(height="500px")
```

---

# Observations and Variables

On the previous slide and in general:

- Each row corresponds to an *observation*
  - In this example that is a State.
- Each column corresponds to a *variable* 
  - In this example that is an attribute of each State.

---

# Histogram: Income

```{r, echo=FALSE, warning=FALSE,fig.align='center'}
States<-readRDS('StateSE.rds')

ggplot(States, aes(x=Income))+geom_histogram(bins = 15)->gp
ggplotly(gp,dynamicTicks = TRUE)%>%
  frameWidget(width="100%",height="40%")
```

---

# Scatter-plot: Income v Mortality

```{r, echo=FALSE,fig.align='center'}
ggplot(States, aes(x=Income, y=LifeExp, label=StateAbb, text=State))+geom_text(size=4)->gp
ggplotly(gp,dynamicTicks = TRUE, tooltip = 'text')%>%
  frameWidget(width="100%",height="50%")
```

---

# 3D Scatter-plot

```{r,  echo=FALSE,eval=TRUE,fig.align='center'}
scatterplot3d(x = States$Income, 
              y = States$LifeExp, 
              z = States$HSGrad, 
              xlab='Income', 
              ylab='LifeExp', 
              zlab='HSGrad', 
              cex.symbols = 2, 
              pch=15)

```

---

# 3D Scatter-plot

Click and drag to rotate

```{r,  testgl, webgl=TRUE,eval=TRUE, cache=TRUE, echo=FALSE,fig.align='center',messages=FALSE,warning=FALSE}
knit_hooks$set(webgl = hook_webgl)
invisible(open3d())
plot3d(x = States$Income, y = States$LifeExp, z = States$HSGrad, xlab='Income', ylab='LifeExp', zlab='HSGrad', size=8,)
rglwidget()
```

---

# Lessons learnt

- With 2 variables we can do a 2-dimensional (2D) scatter plot.
  - This can be interpreted very easily
- With 3 variables we can do a 3D scatter plot
  - This doesn't look great on a flat screen
  - We get more insight by *rotating* the plot
- What about 5 variables? What about 100 variables?

---

# Principal components

- Later on we will cover the method of *principal components*.
--

- This can be used to combine the variables into a single index.
--

- This single index explains most of the variation in the data.
--

- On the next slide we plot the first principal component on a map of the USA.

---

# One PC on a map

```{r, echo=FALSE,eval=TRUE, fig.align='center',fig.width=14,fig.height=10, message=FALSE}

States%>%
  select_if(is.numeric)%>%
  prcomp(scale=T)%>%
  augment(data=States)%>%
  select(-.rownames)%>%
  rename_at(.vars = 8:12,
            function(x){substring(x,8)})->StateSEPC

state_map<-map_data("state")
ggplot(StateSEPC%>%mutate(State=tolower(State)),
       aes(map_id=State))+
  coord_map("albers",lat0=30,lat1=40)+
  geom_map(aes(fill=PC1), map=state_map)+
  geom_text(aes(x=state.center$x,y=state.center$y,label=StateAbb),size=5)+
  expand_limits(x=state_map$long, y=state_map$lat)+
  scale_fill_gradient2()+#n(colors=heat.colors(20))+
  theme_dark()+
  labs(x="",y="")
```

---

# Multidimensional Scaling

- Two states close to one another on the scatterplot had similar levels of income, and life expectancy.
--

- Can we do something similar but for all five variables.
--

- The method of *multidimensional scaling* finds two coordinates so that states close to one another on the scatterplot are close to one another across all five characteristics.

---

#Multidimensional Scaling

```{r, mdsstate, eval=TRUE}
StateSEPC%>%
  ggplot(aes(x=PC1,y=PC2,text=State,label=StateAbb))+
  geom_text(size=4)->gp
ggplotly(gp,dynamicTicks = TRUE, tooltip = 'text')%>%
  frameWidget(width="100%",height="50%")
```

---

# Factor Analysis

- Later on we will attempt to attach possible interpretations to these constructed variables.
--

- This is the objective of factor modelling.
--

- In this context *factor* refers to a latent construct that cannot be directly observed but can be measured via its correlation with observable data.

---

# Cluster Analysis

- Even from the simple analysis so far, it appears that similar states can placed into a small number of groups.
--

- The use of algorithms that achieve this task is known as *cluster analysis*.
--

- It is extremely useful across a number of business disciplines.
--

- On the following slide we group the states into two clusters and present them in different colors.

---

# Cluster Analysis: Example

```{r,echo=FALSE,eval=TRUE, fig.align='center'}
set.seed(1)
States%>%
  select_if(is.numeric)%>%
  scale()%>%
  kmeans(2)%>%
  augment(StateSEPC)%>%
  mutate(Cluster=ifelse(.cluster==.cluster[1],'South','Non-South'))%>%
  select(-.cluster)->StateSEPC
  
ggplot(StateSEPC,aes(x=PC1,y=PC2, label=StateAbb,text=State,color=Cluster))+
  geom_text()->gp
ggplotly(gp,dynamicTicks = TRUE,tooltip = "text")%>%
  frameWidget(width="100%",height="60%")
```

---
class: inverse, middle, center

# A broad understanding of data.

---

# Numerical Data

- So far we looked at *numerical* data 
  - This is also called *metric* data or *ratio* data
- The differences and ratios between values of the variable have some meaningful interpretation.
--

- A state with a mean income of $5000 has twice as much income as a state with a mean income of $2500.

---

# Non-metric data

- *Categorical* (or *nominal*) Data
  - The value of the variable does not measure the *size* of some characteristic.
- *Ordinal* data
  - Different values of the variable measure *more* or *less* of a characteristic but not *how much* more or *how much* less.  

---
# Beer Data

```{r, echo=FALSE, table.align='center', table.width=5}
Beer<-readRDS('Beer.rds')
kable(Beer,format='html')%>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"))%>%
  scroll_box(height="500px")
Beer<-mutate(Beer,Orig=ifelse(origin=='USA','US','Int.'))
```

---

# Questions for you

- How many variables in the Beer dataset?
- Which are metric?
- Which are nominal?
- Which are ordinal? 

---

# Discussion

- Price is an example of a numerical variable.
- Country of Origin is an example of a nominal variable:
  - You can not have more or less *France-ness* or *Mexico-ness*
- Rating is an example of an ordinal variable:
  - A very *good beer* is better than a *good* beer but we do not know how much better.

---

# Cross tab

- A useful tool for exploring non-metric variables is the cross tab.
--

- Cross tabs that are small can be very useful in providing some indication of the relationships between categorical variables. 
--

- Since most Beers in our dataset are from the US, the following cross tab only looks at US beers against beers from all other countries combined. 

---

# Cross Tab: Rating v Origin

International v US
```{r, echo=FALSE}
Beer<-mutate(Beer,Orig=ifelse(origin=='USA','US','Int.'))
kable(table(Beer$rating,Beer$Orig))
```

Is there a relationship between origin and rating?

---

# Using all countries

```{r, echo=FALSE}
kable(table(Beer$rating,Beer$origin))%>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"))%>%
  scroll_box(height="300px")
```

Is it as easy to find a relationship now?

---

#Correspondence Analysis

- Large cross tabulations can be summarised and visualised with a technique known as *Correspondence Analysis*.
--

- This technique is mostly used to visualise the relationship between two variables.
--

- The problem is considered *high-dimensional* since the number of categories rather than the number of variables is large.
--

- On the next slide is the output from correspondence analysis

---

# Correspondence Analysis

```{r, eval=TRUE, echo=FALSE}
table(Beer$rating,Beer$origin)%>%
  ca('symmetric')->caout

caout$rowcoord%>%
  as.data.frame%>%
  rownames_to_column()%>%
  as_tibble%>%
  add_column(Category='Row')->dfr

caout$colcoord%>%
  as.data.frame%>%
  rownames_to_column()%>%
  as_tibble%>%
  add_column(Category='Col')->dfc

cadf<-bind_rows(dfr,dfc)
cadf%>%
  ggplot(aes(x=Dim1,y=Dim2,col=Category,label=rowname))+
  geom_point(size=4)+geom_text_repel(size=7)

```

---

# Other data

- Data comes in even more unusual forms. 
  - The list of your favourite musicians on Spotify
  - The words used in online reviews of hotels
  - A ranking of pairs of products from most similar to most dissimilar
--

- All of these types of data can be analysed using methods covered in the unit.

---
  
# What the unit involves

- The focus is on learning the methods and applying them to real business problems.
--

- To truly understand the methods requires us to learn about two things that students tend to be afraid of:
  - Programming
  - Maths
- This unit has plenty of both!

---

# Why programming is easy

- Programming in this unit uses the *R language*
--

- You will be given lots of support in both lectures and tutorials
--

- Scripts will be provided on Moodle
--

- The best way to learn R is to: 
  - Practice 
  - Make mistakes 
  - Use the Internet!
- In previous years students have found learning R extremely useful.

---

# Why math is easy

- People who hate math tend to hate memorising formulas
--

- So do I...
--

- The emphasis in this unit is 
 - Good intuition 
 - Recognising patterns 
 - Using your imagination.
- Most of the math we cover has a *geometric* interpretation which is great for visual learners.

---

# How the unit is delivered

- There is no textbook.  The lecture notes and tutorials make up the content of the unit.
--

- Lectures are recorded.
--

- However, you will be asked to work in the middle of lectures while I walk around answering questions.
--

- Install R on laptops and use in class.
--

- Forget to bring a laptop? Share with a friend.
--

- Tutorials will be a mix of reinforcing concepts from the lecture and using R.

---

# How to succeed in this unit

- If your only motivation is to pass the unit
  - Follow lectures and attend tutorials.
  - Submit assignments on time.
  - Use common sense.
- If your motivation is to get a high distinction
  - Do all the above.
  - Attempt the practice questions between lectures and tutorials.
  - Try to teach yourselves how to do things in R. Treat it like a puzzle.
  - Understand matrix algebra.

